---
title: "BUILDING AN ASSESSMENT-READY CONDUCT DATA MODEL"
subtitle: "Answering Strategic Questions About Student Conduct Through Structured Data"
description: "A Guide for Senior Leaders, Conduct Professionals, and Data Partners"
author: ""
date: last-modified
---
```{=latex}
\pagestyle{empty}
\clearpage
\pagestyle{scrheadings}
\pagenumbering{arabic}
\setcounter{page}{1}
```

\newpage
# Executive Summary

**The Problem**

Student conduct data from Maxient captures what happens inside the case management system, but meaningful analysis often requires context that lives outside it: campus and residence hall populations, academic college affiliations, staff assignments, and institutional calendar structures. Without a way to connect these sources, practitioners are left with incomplete pictures and exports that are difficult to compare consistently across years.

**The Solution**

This project transforms raw Maxient exports into a star schema data model that Power BI can use reliably, one where filters work as expected, numbers stay stable, and stakeholders can trust what they're seeing.

These dashboards are designed to support aggregate analysis and strategic conversation. They are not a substitute for case-level review, professional judgment, or due process.

**What This Enables**

Structuring data this way makes it possible to:

- Compare trends across academic years without manual data cleanup
- Filter by hall, term, case type, or staff assignment and get consistent counts
- Calculate violation rates based on population for fair comparisons across buildings
- Track how long cases take to resolve, adjusted for breaks and holidays
- Share dashboards more broadly because student identifiers have been removed
- Use aggregate trends and rates to inform policy discussions, resource planning, and prevention strategies grounded in professional judgment and institutional context

\newpage
**Who Should Read What**

| **If you are...** | **Focus on...** | **You can skip...** |
|---------------|-------------|-----------------|
| Senior Leadership | Executive Summary and Section 4 (Common Questions) | Everything else |
| Unit Directors, Deans | Executive Summary, Section 4, callout boxes throughout | Code blocks, technical details |
| Conduct Professionals | Sections 1–4, table descriptions in Section 5 | Code blocks |
| Analysts and IT Professionals | Everything, especially code blocks and Section 6 | Nothing |

**Summary:** Senior leaders can focus on the Executive Summary; directors/deans should add Section 4; conduct pros should read Sections 1–4; analysts/IT should read the full document.

\newpage

# Accessibility Statement

This document has been designed with accessibility in mind and strives to meet Web Content Accessibility Guidelines (WCAG) 2.1 Level AA standards. However, due to current limitations in LaTeX-based PDF generation technology, some accessibility features require post-processing:

**Known Limitations:**

- **PDF Tagging**: LaTeX does not natively generate tagged PDFs with full semantic structure. This document will undergo automatic tagging in Adobe Acrobat Pro to ensure proper document structure for assistive technologies.

- **Table Headers**: While all tables in this document use proper header rows, LaTeX-generated PDFs may not automatically mark them as header cells.

- **Code Blocks**: R code chunks are rendered as syntax-highlighted text but may be interpreted by PDF readers as images or figures. During accessibility review, these are marked as decorative. Descriptive text accompanies each code section to ensure the methodology is accessible regardless of code readability.

**Commitment to Accessibility:**

The author is committed to making this document accessible. If you encounter any accessibility barriers or need this document in an alternative format, please contact:

- Joshua L. Moermond
- Email: [moermondsahe@gmail.com](mailto:moermondsahe@gmail.com)

HTML and Quarto markdown versions of this document are available in the project repository on [Github](https://github.com/stdntaffrsanalyst/conduct-data-model) and may provide better compatibility with some assistive technologies.

\newpage

# Introduction

Student conduct systems like Maxient are built to help staff manage cases: documenting incidents, tracking outcomes, and ensuring follow-up. Maxient does this exceptionally well — it is the dominant platform in the field for good reason. But case management and strategic analysis are genuinely different disciplines, and a system optimized for one is not automatically suited for the other. The same data that supports day-to-day case management becomes difficult to use when the questions get bigger: comparing trends across years, connecting to institutional context, calculating rates relative to population, or sharing findings with campus partners who need to understand what's happening but don't work in the system. Those questions require a different structure.

Three specific gaps make raw Maxient exports difficult to use for strategic analysis. The first is data quality over time. Maxient's data integrity depends on consistent discipline from everyone entering records — and in practice, that discipline is uneven. Institutional configurations also change over time, so similar behavior may be recorded under different charge names across years, and the same location might be spelled three different ways. And a single export row contains the incident, the charges, the outcome, and the sanctions all at once, making it hard to count any one thing cleanly.

The second gap is analytical scope. Maxient can tell you how many alcohol incidents occurred this year. It cannot tell you whether the rate is increasing relative to residential population, which halls are disproportionately affected, or how case timelines vary across staff assignments or employee groups. Those questions require connecting Maxient data to institutional context that lives outside the system: residence hall populations, academic college affiliations, staff assignments, and calendar structures. Without that connection, you can count what happened — but you can't fairly compare across buildings, years, or people.

The third gap is access. Meaningful analysis — tracking repeat involvement, linking records across years, connecting cases to student characteristics — requires including student and case identifiers in the working dataset. This creates a real constraint: a file built for analysis isn't safe to share broadly, even when the patterns it reveals would be appropriate for housing leadership, senior administrators, or accreditation reviewers to see.

This project addresses all three gaps. It combines Maxient exports with supplementary institutional data, transforms the result into a star schema data model designed for Power BI, and anonymizes identifiers so that findings can be shared broadly without exposing the students behind them.

## What Is a Star Schema?

Think of a star schema as an organized filing system where related information connects like a hub-and-spoke wheel.

At the center (the hub) are tables recording events and measurements: how many incidents occurred, how long cases took to resolve, which sanctions were assigned. These are called 'fact tables.' Each row in a fact table represents one one unit of measurement — an event, a duration, or a calculated value — at a specific level of detail.

Radiating outward (the spokes) are tables storing the descriptive context used for filtering and grouping: when something occurred, where, who was involved, what type of case. These are called 'dimension tables.' Every dimension table connects to one or more fact tables through a shared identifier — this connection is what allows a filter on 'residence hall' to correctly limit which incidents are counted.

Each fact table is built around a specific grain — the precise unit of measurement one row represents. For example, one row in FactViolation represents one charge in one case. Keeping the grain consistent within a table is what makes counts reliable.

Why does this matter? When these two types of information are mixed together in one giant spreadsheet, dashboards can give inconsistent answers. Filtering by one thing accidentally changes the count of something else. Numbers shift depending on how you slice the data. People stop trusting the reports.

A star schema prevents this by keeping "what happened" and "context" in separate but connected tables. The result: filters work as expected, counts stay stable, and stakeholders can trust what they're seeing.

::: {.callout-tip}
## *Why This Matters*

**For Decision-Makers:** Dashboards become trustworthy. When you filter by residence hall or academic year, the numbers won't mysteriously change based on what else is on the page.

**For Analysts:** This is the structure Microsoft recommends for Power BI, and is based on dimensional modeling principles widely used in data warehousing. It improves query performance and ensures DAX measures calculate correctly across filter contexts. A full entity-relationship diagram of the model is available in the GitHub repository.
:::

\newpage
The ultimate goal is to support better decisions in student conduct work:

- **Staffing**: Where is the workload concentrated? Do we need more support in certain areas or at certain times?
- **Policy**: Are our policies producing the outcomes we expect? What happens after we make changes?
- **Equity**: Are patterns consistent across student populations and environments, or are there disparities worth investigating?
- **Trends**: How is this year different from last year? Are changes meaningful or just noise?

## How to Read This Document

### About the Code Blocks

Throughout this document, you'll see gray boxes containing R code. These show *how* each table is created behind the scenes. They're included for transparency and so the process can be repeated exactly the same way each time.

If you're not familiar with R or data analytics, you can skip the code entirely and still understand what the model does and why it matters.

::: {.callout-tip}
## *Why Include Code?*

**For Decision-Makers:** You don't need to read it, but knowing it exists means this process doesn't depend on one person's memory. If Joshua wins the lottery and moves to Tahiti, someone else can pick this up.

**For Analysts:** The code is the documentation. You can trace exactly what happens to each field, verify the logic, and adapt it for new requirements.
:::

## Scope and Non-Goals

This project is designed to support program-level assessment, strategic decision-making, and institutional learning in student conduct and student affairs contexts. It focuses on transforming operational case-management data into a structure that supports reliable aggregation, trend analysis, and comparison over time.

This model is not intended to:

- Replace case-level review, professional judgment, or due process.
- Establish causal relationships between policies, interventions, and outcomes.
- Serve as a substitute for individualized performance evaluation. Aggregate metrics like average case resolution time can inform supervision and identify patterns worth discussing, but they do not capture the full context of any individual's work.
- Generalize directly to other institutions without local adaptation.

While the transformation pipeline produces stable and reproducible metrics, interpretation of those metrics must always be grounded in institutional context, policy intent, and professional practice.

\newpage

# Definitions & Reporting Concepts

This section defines key terms used throughout the document. You don't need to memorize these; they're here as a reference when you encounter unfamiliar terminology.

## Grain

**Grain** answers the question: *What does one row in this table represent?*

This single concept prevents most reporting disagreements. When two reports 
show different numbers, the first question should always be: *What are we 
counting?*

Think of it like a receipt. A grocery store receipt has one line per item 
purchased — that's the grain. If you count lines on your receipt, you get the 
number of items, not the number of shopping trips or the number of stores.

**Row Meaning Across Core Conduct Tables**

The table below explains what a single row represents in each major conduct 
data table and what counting rows will tell you.

| **Table** | **What one row represents** | **If you count rows, you get...** |
|-------|------------------------|------------------------------|
| FactIncident | One FILE_ID | Number of involved parties — use distinct CASE_NUMBER for incident counts |
| FactViolation | One alleged charge | Number of charges filed |
| FactSanction | One sanction applied | Number of sanctions given |
| DimStudent | One unique student | Number of students in the system |

**A note on FILE_ID and incidents:** FILE_ID is Maxient's identifier for an 
individual involved party within an incident. An incident involving two people produces two FILE_IDs and therefore two rows in FactIncident. To count incidents rather than involved parties, always count distinct CASE_NUMBER values.

**Why this matters:** A single incident might involve multiple parties, 
multiple violations, and result in multiple sanctions. A student involved in 
three incidents appears in three rows of FactIncident but only one row in 
DimStudent.

If someone asks "How many students were involved in incidents last year?" and 
you count rows in FactIncident, you'll overcount — both because students can 
have multiple incidents and because incidents can have multiple parties. 
Count distinct SID values instead, or use DimStudent.

**In summary:**

- Counting distinct CASE_NUMBERs in FactIncident answers "How many incidents 
  occurred?"
- Counting rows in FactIncident answers "How many involved parties were 
  recorded?"
- Counting rows in FactViolation answers "How many charges were filed?"
- Counting rows in FactSanction answers "How many sanctions were issued?"
- Counting rows in DimStudent answers "How many unique students are 
  represented?"\

::: {.callout-tip}
## *The Fastest Way to Resolve Number Disagreements*

**For Decision-Makers:** When two reports show different numbers, ask: "What are we counting—cases, incidents, violations, sanctions, or students?" The answer usually explains the discrepancy.\

**For Analysts:** Always document the grain of any metric you create. "Incident count" and "student count" can differ by 30% or more depending on repeat involvement rates.
:::

## Fact Tables vs. Dimension Tables

### Fact Tables: What Happened

A fact table stores **events and measures**—the things you want to count or analyze. Each row represents something that occurred.

The table below summarizes what each fact table represents and the types of questions it is designed to answer.

\newpage
| **Fact Table** | **What each row represents** | **Example questions it answers** |
|------------|-------------------------|------------------------------|
| FactIncident | One incident | "How many incidents occurred in Fall 2024?" |
| FactViolation | One alleged charge | "What's the not-responsible rate for alcohol violations?" |
| FactSanction | One sanction applied | "How often do we assign educational programs vs. probation?" |
| FactTimeline | Timing for one case | "How long do cases take from report to resolution?" |
| FactRecidivism | Repeat involvement summary by year | "What percentage of students have multiple cases within an academic year?" |
| FactCohortRecidivism | One student (by first responsible finding) | "Of students who first got in trouble in AY2223, how many ever got in trouble again?" |
| FactHousingCensus | Violations + population for one hall/term | "Which hall has the highest violation rate per 100 residents?" |
| FactDeadline | One case deadline | "How many cases are currently overdue?" |

**In summary:**

- FactIncident counts incidents.
- FactViolation counts charges.
- FactSanction counts sanctions assigned.
- FactTimeline measures process duration.
- FactRecidivism and FactCohortRecidivism measure repeat involvement.
- FactHousingCensus enables rate calculations using population.
- FactDeadline tracks open and overdue cases.

### Dimension Tables: Context About What Happened

A dimension table stores **descriptive information** that helps you filter and interpret events.

The table below summarizes what each dimension table represents and how it is typically used when filtering or grouping conduct data.

| **Dimension Table** | **What it describes** | **Example use** |
|-----------------|-------------------|-------------|
| DimDate | Calendar and academic time | Filter by semester, compare academic years |
| DimAcademicYear | Academic year boundaries | Ensure AY2324 means Aug 2023–Jul 2024 everywhere |
| DimAcademicTerm | Fall/Spring/Summer terms | Compare Fall-to-Fall trends |
| DimAcademicPlans | Academic plan codes and major names | Translate coded major fields to human-readable values for college and career-level analysis |
| DimHousing | Residence hall attributes | Filter by building type or ownership |
| DimHousingYear | Hall attributes that change yearly | Track capacity changes over time |
| DimStudent | Student demographics/academics | Disaggregate by classification, college, ethnicity |
| DimCase | Case-level attributes | Filter by case type, status, assigned officer |
| DimSanction | Sanction categories and severity | Group sanctions by type (educational, restrictive, etc.) |
| DimCollege | Academic colleges | Filter or group by college |
| DimHearingOfficer | Staff assignments | Analyze workload by officer or department |
| DimViolation | Charges metadata (categories, severity, etc.) | Group charges by type and severity |

**In summary:**

- DimDate, DimAcademicYear, and DimAcademicTerm define time structures for filtering and comparing periods.
- DimHousing and DimHousingYear describe residential context, including attributes that remain stable versus those that change annually.
- DimStudent, DimCollege, and DimHearingOfficer describe people and organizational context for disaggregation and workload analysis.
- DimCase, DimSanction, and DimViolation provide case and outcome metadata used to group, filter, and interpret incident details.\

::: {.callout-tip}
## *Why Separate Facts from Context?*

**For Decision-Makers:** This separation means that when you filter by residence hall or semester, you change *which* incidents are included—but you don't change *what* "incident" means. The definition stays stable.\

**For Analysts:** In Power BI terms, this creates clean one-to-many relationships. Dimensions filter facts, not the reverse. This prevents circular dependencies and ambiguous filter propagation.
:::

## Counts vs. Rates

A **count** is the total number of events: *"40 violations in Calhoun Hall."*

A **rate** compares a count to a population: *"5 violations per 100 residents."*

Counts describe volume, not severity, intent, or community climate. They are most useful as a starting point for inquiry, not as a standalone measure of activity or concern within a location or population.

**When to use each:**

| **Question Type** | **Use** | **Example** |
|---------------|-----|---------|
| "How much work is this?" | Count | "We had 200 cases this semester—do we need more staff?" |
| "Where should we focus prevention?" | Rate | "Building A has 12 violations per 100 residents vs. Building B's 4 per 100" |

\newpage
**Example:** 

| **Hall** | **Violations** | **Residents** | **Rate per 100** |
|------|------------|-----------|--------------|
| Building A | 50 | 800 | 6.25 |
| Building B | 25 | 200 | 12.50 |

Building A has twice as many violations, but Building B's *rate* is twice as high. Which number matters depends on your question:

- For staffing: Building A (workload), the count matters.
- For targeting prevention efforts (where is behavior concentrated?), the rate matters.\

::: {.callout-tip}
## *Fair Comparisons Require Rates*

**For Decision-Makers:** When comparing halls, terms, or populations of different sizes, always ask for the rate, not just the count. A hall with higher counts may simply have more residents.\

**For Analysts:** FactHousingCensus pre-calculates violation rates per 100 residents by hall and term, joining violation counts with census data.
:::

## Time Intelligence: Academic vs. Calendar Time

Student Affairs work follows the academic calendar, not the calendar year. "Last year" usually means "last academic year" (August to July), not "last calendar year" (January to December).

This model includes a dedicated date table (DimDate) that ensures:

- **"AY2324" means the same thing everywhere**: August 1, 2023 through July 31, 2024
- **Semester comparisons are consistent**: Fall is always August–December; Spring is always January–April
- **Year-over-year trends align properly**: Comparing Fall 2023 to Fall 2022, not December 2023 to December 2022

The related tables DimAcademicYear and DimAcademicTerm provide additional structure for filtering and grouping by academic periods.\

::: {.callout-tip}
## *Why Academic Time Matters*

**For Decision-Makers:** Reports that use calendar years can obscure patterns. A spike in "January 2024" might be the end of Fall 2023's cases being resolved, not a Spring 2024 trend.\

**For Analysts:** DimDate includes an AcademicTermKey that links to DimAcademicTerm, enabling correct semester-level aggregation. The AY_Order field ensures academic years sort correctly (AY2223 before AY2324).
:::

\newpage

# Methodology

This section describes *how* the transformation works at a conceptual level. If you want to understand the approach without reading code, this section is for you. The actual implementation details appear in Section 5.

## Overview
### Six Steps from Messy to Usable Conduct Data

1. **Combine**
    Multiple Maxient exports are merged into a single, unified dataset so that trends can be analyzed consistently across academic years.

2. **Standardize**  
   Inconsistent values -- charge names, location spellings, ethnicity codes, role labels -- are normalized to single standard representations. this step also addresses configuration drift. For example, when the same behavior has been recorded under different charge names over time, normalization ensures those records are treated as equivalent for analysis.
   
3. **Enrich**
   External data sources are joined to the Maxient export to add analytical context that doesn't exist in the case management system: building characteristics and resident counts from housing, staff names and departments from hearing officer assignment tables, and analyst-defined classifications for sanctions and violations including category labels, severity levels, and restrictive flags. These enrichment fields are not Maxient-native — they represent deliberate analytical decisions that must be documented and maintained alongside the pipeline.

4. **Add Academic Time**  
   Raw date fields are transformed into academic-year, semester, week, and day-of-week variables so that reporting aligns with how Student Affairs work is organized and evaluated. A dedicated date table is built to enable reliable time-based filtering in Power BI.

5. **Calculate Process Metrics**  
   Key timeline metrics — days from report to case creation, case creation to resolution, and report to resolution — are computed and adjusted to exclude institutional breaks and holidays so that metrics reflect working time rather than calendar days. Timeline logic differs by case type and must be applied accordingly.

6. **Restructure and Protect Privacy**  
   Direct identifiers are anonymized using a secure, irreversible hashing method. The wide Maxient export — one row per case with charges and sanctions spread across columns — is reshaped into fact and dimension tables, each with a defined grain, so that dashboards can be shared broadly and Power BI can query the data reliably.

Each step addresses a specific barrier that makes raw Maxient data difficult to use for strategic analysis.

## Step 1: Combine Multiple Years

**The Problem:** Longitudinal analysis requires data from multiple academic years to be combined into a single consistent dataset. Maxient exports are typically pulled in batches, and those files may have slightly different column structures depending on when they were created or what system updates occurred in the interim.

**The Solution:** The pipeline reads all export files and merges them into a single data set, handling any column differences automatically.

**Why This Matters:** Longitudinal analysis—comparing this year to last year, or tracking five-year trends—requires data to be in one place. Without this step, every trend analysis would require manual file merging.\

::: {.callout-tip}
## *Why Combine Data?*

**For Decision-Makers:** You can now answer questions like "How has incident volume changed over five years?" without anyone manually copying and pasting spreadsheets together.\

**For Analysts:** The pipeline uses `bind_rows()` which handles column mismatches gracefully—columns present in some files but not others become NA values rather than causing errors.
:::

## Step 2: Standardize Key Fields

**The Problem:** The same information gets recorded differently over time and across users:

| **Field** | **Variations Found** | **Standardized To** |
|-------|------------------|-----------------|
| Location | "CALHOUN," "Calhoun Hall," "Calhoun" | "Calhoun Hall" |
| Location | "Stratford Hts Bld 2," "Stratford Heights Bldg 2" | "Stratford Heights" |
| Ethnicity | "BLACK," "Black NotHisp," "African-American" | "Black" |
| Role | "Alleged/Respondent," "Respondent" | "Respondent" |
| Sport | "M-Football," "Football," "Men's Football" | "Men's Football" |

**The Solution:** The pipeline applies consistent mappings that convert all variations to a single standard value. It also:

- Converts date fields to proper date format (not text)
- Replaces blank values with "Not Reported" so they show up in filters instead of disappearing
- Recodes obsolete categories (e.g., "Pre-Junior" → "Sophomore")

**Why This Matters:** If "Calhoun Hall" and "CALHOUN" are treated as different places, your counts are wrong. Standardization ensures that when you filter by a location, you capture all relevant records.

Standardization improves consistency for analysis, but it does not eliminate professional judgment or contextual nuance in how cases are documented and interpreted.\

::: {.callout-tip}
## *Why Standardization Protects Your Analysis*

**For Decision-Makers:** You can trust that "Calhoun Hall" in the dashboard means *all* Calhoun Hall incidents, not just the ones that happened to be spelled a certain way.\

**For Analysts:** Standardization happens once, upstream, in the pipeline. This means every downstream table and dashboard inherits clean values without needing its own cleanup logic. The mappings are documented in `config.R` and can be updated as new variations appear.
:::

### A Note on Configuration Drift and Historical Data Integrity

Standardization corrects inconsistencies in the current export — but it cannot fully resolve a deeper structural challenge: configuration changes accumulate over time, and Maxient captures data at the moment of case creation.

**Configuration drift** occurs when charge names, location labels, or other categorical values change across academic years. "Alcohol — Underage Possession," "Alcohol — Distribution," and "Alcohol" may all represent the same policy violation but return as separate categories in any report that spans the years they were in use. Standardization mappings address this directly — but those mappings must be actively maintained as new variations appear.

**Historical data integrity** is a related but distinct problem. When your institution corrects a configuration error — for example, fixing how major codes are passed from the SIS to Maxient — the correction only applies to cases created after the fix. Historical records retain the values that existed at case creation. The result is a split dataset: recent cases carry readable values while older cases carry codes or outdated labels, and no configuration change can retroactively clean them.

**Temporal inconsistency** refers to the consequence: the same field means different things depending on when a record was created. Analysis that spans multiple academic years must account for this. The standardization mappings in config.R are the primary mechanism for managing this — but they require ongoing professional judgment about what to map, what to leave as-is, and how to document decisions for future analysts.

::: {.callout-tip}
## *What This Means for Trend Analysis*

**For Decision-Makers:** Multi-year trend comparisons are only as reliable as the standardization logic behind them. If charge names changed three years ago and the mapping wasn't updated, historical comparisons for that violation type may be undercounting or overcounting. Always ask: has this field been consistent across the years being compared?\

**For Analysts:** Review and update `config.R` standardization mappings at the start of each academic year or whenever Maxient configuration changes are made. Document the reason for each mapping and the date it was added. This is institutional knowledge that must survive staff turnover.
:::

## Step 3: Enrich with External Context

**The Problem:** Maxient captures what happened inside a case. It does not know how many students live in a building, what department a hearing officer works in, or whether a sanction is considered restrictive under your institution's framework. Without this context, many of the most important strategic questions — violation rates by building population, workload by staff assignment, sanction mix by educational versus restrictive category — cannot be answered.

**The Solution:** The pipeline joins external data sources to the Maxient export and adds analyst-defined classification fields:

| **Source** | **What It Adds** | **Why It Matters** |
|--------|--------------|----------------|
| Housing characteristics table | Building type, ownership model, room types | Enables filtering and comparison by residential context |
| Housing census data | Resident counts by building and term | Required for rate calculations (violations per 100 residents) |
| Hearing officer assignment table | Staff names, departments, position titles | Enables workload analysis and disaggregation by staff |
| DimSanction classification | CATEGORY_LABEL, SANCTION_CATEGORY, RESTRICTIVE_FLAG, SEVERITY_LEVEL, SEVERITY_LABEL | Groups sanctions for educational vs. restrictive analysis |
| DimViolation classification | CATEGORY, IS_ALCOHOL, IS_DRUG, IS_VIOLENCE, SEVERITY_LEVEL, SEVERITY_LABEL | Groups charges for pattern analysis and severity comparison |

**An important distinction:** The Maxient-native fields in DimSanction are SANCTION and DESCRIPTIVE_NAME. The fields CATEGORY_LABEL, SANCTION_CATEGORY, RESTRICTIVE_FLAG, SEVERITY_LEVEL, and SEVERITY_LABEL are analyst-defined — they represent deliberate classification decisions that are not generated by Maxient. The same applies to DimViolation: VIOLATION and POLICY are Maxient-native; the category, severity, and flag fields are analyst-defined. These decisions should be documented, reviewed periodically, and updated when institutional policy or practice changes.

**Why This Matters:** Raw conduct data tells you what happened. Enriched conduct data tells you what it means in institutional context. The difference between "50 violations in Calhoun Hall" and "6.25 violations per 100 residents in Calhoun Hall" requires enrichment. The difference between "a suspension was issued" and "a restrictive sanction was issued" requires analyst classification.

::: {.callout-tip}
## *Enrichment vs. Raw Data*

**For Decision-Makers:** When a dashboard shows violation rates, sanction severity, or staff workload, those numbers depend on enrichment data — housing census figures, staff assignment tables, and analyst classifications — that lives outside Maxient. If those sources are outdated or incorrect, the derived metrics will be too. Enrichment data should be reviewed and updated at the start of each academic year.\

**For Analysts:** Enrichment joins use left_join() to preserve all Maxient records even when a match isn't found in the external table. Unmatched records receive "Not Reported" values. Review join match rates after each pipeline run — a sudden drop in matches often indicates a configuration change in how Maxient or the SIS passes field values.
:::

## Step 4: Add Academic Time Structure

**The Problem:** Maxient stores dates as calendar dates, but Student Affairs work is organized around the academic calendar. Questions like "How does this semester compare to last semester?" or "What happened during Fall 2023?" require academic time fields that don't exist in the raw data.

**The Solution:** The pipeline derives academic time fields from incident dates:

| **Derived Field** | **What It Captures** | **Example** |
|---------------|------------------|---------|
| ACADEMIC_YEAR | Aug 1 – Jul 31 cycle | "AY2324" (Aug 2023 – Jul 2024) |
| Semester | Term within the year | "FA" (Fall), "SP" (Spring), "SU" (Summer) |
| WEEK_NUM | ISO week number | Week 1–52 for week-over-week patterns |
| incident_month | Month, ordered Aug–Jul | Allows academic-year month trends |
| incident_wday | Day of week | For day-of-week patterns |

**Why This Matters:** A report showing "January 2024 had high case volume" might be misleading—that could be Fall 2023 cases reaching resolution, not a Spring 2024 trend. Academic time fields let you interpret patterns in context.\

::: {.callout-tip}
## *Why Academic Time Changes Everything*

**For Decision-Makers:** You can now ask "How does Fall 2024 compare to Fall 2023?" and get a meaningful answer. Calendar-year reports often obscure the patterns that matter for academic operations.\

**For Analysts:** The DimDate table provides a complete date spine with pre-calculated academic year, semester, and term keys. Fact tables join to DimDate on incident date, enabling time intelligence in Power BI without complex DAX calculations.
:::

## Step 5: Calculate Process Metrics

**The Problem:** Understanding workflow efficiency requires knowing how long cases take to move through the process. But simply subtracting dates can produce misleading results:

- A case reported December 15 and resolved January 15 looks like 31 days—but the office was closed for winter break for 20 of those days.
- A case might show negative days if dates were entered incorrectly.

**The Solution:** The pipeline calculates three timing metrics:

| **Metric** | **What It Measures** | **Calculation** |
|--------|------------------|-------------|
| reported_to_case | Intake/triage speed | Days from report date to case creation date |
| case_to_resolution | Adjudication speed | Days from case creation to hearing/resolution date |
| reported_to_resolution | Total process time | Days from report to resolution |

Each calculation:

- **Subtracts institutional pause periods** (breaks, holidays) so a case that spans winter break isn't penalized for closed days

- **Floors negative values to zero** so data entry errors don't produce impossible results

**Why This Matters:** Timing metrics help identify where delays occur. Is the bottleneck at intake (getting cases created) or adjudication (getting cases resolved)? Are certain case types systematically slower?\

::: {.callout-tip}
## *How to Interpret Timing Metrics*

**For Decision-Makers:** Longer timelines aren't automatically "bad." They might reflect case complexity, due process requirements, student availability, or staffing patterns. Use timing data as a *signal to investigate*, not a performance score.\

**For Analysts:** Pause periods are defined in `config.R` as `OMIT_RANGES`. The `compute_overlaps()` function calculates how many days of each case's timeline fall within these ranges and subtracts them. Warning Letter cases are excluded from adjustment since they resolve immediately.
:::

## Step 6: Restructure and Protect Privacy

**The Problem:** Raw Maxient exports contain student IDs, case numbers, and other identifiable information. Sharing dashboards broadly—with housing partners, leadership, or for accreditation—creates privacy risk. Additionally, the wide format of Maxient exports (one row per case with many columns) doesn't work well for dashboard tools.  

**The Solution:** The pipeline does two things: it protects privacy and organizes data for dashboard-ready reporting.

### Privacy Protection

Identifiable fields (case numbers, file IDs, student IDs) are scrambled using a secure code. The scrambled values:

- **Look like random text** (e.g., "a3f7c9e2b1d8...")
- **Are consistent** (the same student ID always produces the same scrambled value, so records can still be linked across tables)
- **Cannot be reversed** (you can't work backward from the scrambled value to find the real ID)

This means dashboards can show patterns involving individual students (like repeat involvement) without exposing who those students are.

### Star Schema Organization

The wide Maxient export is split into separate fact and dimension tables:

**Before (one wide row per case):**
```
FILE_ID | Student_Name | DOB | Hall | Charge_1 | Charge_2 | Finding_1 | ...
```

**After (separate tables linked by keys):**
```
FactIncident: FILE_ID | INCIDENT_DATE | INCIDENT_LOCATION | ...
FactViolation: FILE_ID | VIOLATION | OUTCOME | ...
FactSanction: FILE_ID | Sanction | ...
DimStudent: SID | ETHNICITY | CLASSIFICATION | ...
DimHousing: INCIDENT_LOCATION | Housing_Type | ...
```

This separation means:

- Each table has a clear grain (what one row represents)
- Descriptive information isn't duplicated across thousands of rows
- Power BI can filter efficiently because relationships are explicit\

::: {.callout-tip}
## *Why This Enables Broader Access*

**For Decision-Makers:** With identifiable information removed, dashboards can be shared with more stakeholders, including housing partners, senior leadership, and accreditation reviewers, without FERPA concerns. Patterns are visible; individuals are not.\

**For Analysts:** The scrambling uses HMAC-SHA256 with a locally stored key. The same key must be used each time to ensure consistency across refreshes. The key file (`pepper.bin`) should be protected and backed up but never shared or committed to version control.
:::

## Data Sources

### Primary Source: Maxient Case Exports

The pipeline ingests Maxient case exports covering multiple academic years. These exports are generated from Maxient's reporting tools and contain:

- Case and incident information
- Student demographic data (as recorded in Maxient)
- Charges and findings
- Sanctions applied
- Assignment and resolution information

Multiple exports are required because:

1. Maxient is a web-based tool and doesn't perform well with large exports
2. Historical data may need to be re-exported after system updates
3. Different date ranges may have been exported at different times

### Supplementary Sources: Lookup Tables

Several external lookup tables enrich the Maxient data:

| **File** | **Purpose** | **Source** |
|------|---------|--------|
| `DimAcademicPlans.csv` | Maps major codes to college names | Registrar |
| `hearing_officers.csv` | Maps assignment codes to staff names/departments | SCCS |
| `DimHousing.csv` | Housing attributes (type, ownership, room style) | University Housing |
| `DimHousingYear.csv` | Year-specific housing attributes | University Housing |
| `DimSanction.csv` | Sanction categories and severity classifications | SCCS |
| `housing_census.csv` | Resident counts by building and term | University Housing |
| `DimCharge.csv` | Charge categories, severity, and flags | SCCS |
\
These lookup tables allow the dashboard to answer questions that Maxient data alone cannot—like "What's the violation rate per 100 residents?" or "How do patterns differ by academic college?"\

::: {.callout-tip}
## *Why Supplementary Data Matters*

**For Decision-Makers:** Raw conduct data tells you *what happened*. Adding context (like building populations or college affiliations) tells you *what it means*. A hall with 50 violations could be a problem or unremarkable depending on how many students live there.\

**For Analysts:** Lookup tables should be updated annually or when organizational changes occur (new buildings, renamed colleges, staff turnover). The pipeline will still run with outdated lookups, but unmatched values will appear as "Not Reported" in dimensions.
:::

## Output: What the Pipeline Produces

The final output is an Excel workbook containing all dimension and fact tables as separate worksheets. Each worksheet represents a single table in a star-schema model, preserving consistent row-level meaning across academic years. This workbook is designed for direct import into Power BI.

In Power BI, the workbook is imported using Get Data → Excel. Relationships between tables are established using shared key fields (e.g., FILE_ID, SID, INCIDENT_DATE, INCIDENT_LOCATION). Because the tables are already structured for analysis, there should be little to no need for additional structural data transformation within Power Query.

The resulting data model supports dashboards for:

- Volume and trends: Incident counts over time, by location, and by type
- Workload analysis: Cases per officer and timeline distributions
- Outcome patterns: Finding rates by violation type and sanction frequencies
- Equity review: Disaggregation by student demographics
- Housing context: Violation rates per 100 residents by hall
- Process monitoring: Overdue cases and timeline bottlenecks

This approach separates data preparation from visualization, helping ensure that reported numbers remain stable, reproducible, and transparent across reporting cycles.

\newpage

# Data Quality and Reporting Limitations

As with any assessment effort based on administrative data, the outputs of this model reflect both the strengths and limitations of the underlying source systems and reporting practices.

Key considerations include:

1. **Reporting Bias**\
  Conduct data reflects what is reported and documented, not all behavior that occurs. Changes in reporting practices, staff training, or institutional priorities may affect trends independent of student behavior.

2. **Charging Discretion**\
  Violations and charges reflect professional judgment at the time of case intake. Differences in charging practices across staff, time periods, or policy regimes may influence counts and rates.

3. **Historical Inconsistency**\
  Maxient exports may vary across years due to system updates, form changes, or evolving definitions. The pipeline standardizes these differences where possible, but historical context remains important.

4. **Missing or Evolving Data**\
  Some fields may be incomplete, inconsistently populated, or newly introduced. Explicit “Not Reported” values are used to preserve transparency rather than silently excluding records.

5. **Small-Number Instability**\
  Disaggregated analyses (e.g., by specific buildings, demographics, or niche case types) may be sensitive to small counts. Apparent changes may reflect random variation rather than meaningful shifts.

Understanding these limitations is part of using this model well. It is designed as a decision-support tool, interpreted alongside professional expertise, qualitative context, and institutional knowledge, not as a replacement for any of them.

\newpage

# Strategic Questions This Model Can Answer

This section translates the data model into the strategic questions it is designed to answer. If you are wondering what you can actually do with this data, this section is for you.

The questions are organized by category. For each category, you will find example questions, which tables provide the answers, and guidance on interpretation. These categories align with the six strategic question types introduced in session presentations and workshops.

## Quick Reference: Which Table Answers Which Strategic Question?

| **If you're asking about...** | **Start with...** | **Grain** |
|---------------------------|---------------|--------------------------------|
| How many incidents occurred | FactIncident | One incident record |
| Violation types and outcomes | FactViolation | One alleged charge |
| Sanctions and interventions | FactSanction | One sanction applied |
| How long cases take | FactTimeline | One case's timing metrics |
| Repeat student involvement | FactRecidivism | Summary by academic year |
| Violations adjusted for population | FactHousingCensus | One hall in one term |
| Overdue or pending cases | FactDeadline | One case deadline |
| Student characteristics | DimStudent | One unique student |
| Building attributes | DimHousing / DimHousingYear | One building (or building-year) |
| Staff workload | DimCase + DimHearingOfficer | One case |
| Academic plan and major context | DimAcademicPlans | One academic plan code |
| College or academic unit context | DimCollege | One academic college |

Two additional tables support case-level filtering and operational analysis but are not typically the starting point for dashboard questions:

- **FactTag / DimTag**: Flag cases with operationally significant conditions (e.g., dismissed cases, hospital transports). Useful for excluding or isolating specific case types when calculating other metrics.
- **DimAcademicPlans**: Provides the full translation of academic plan codes to major names, career levels, and college affiliations. In most dashboard contexts this context is accessed through DimCollege or DimStudent rather than directly.

## Volume and Trend

These questions help you understand *how much* is happening and *whether patterns are changing*.

**Example Questions:**

- How many incidents occurred this academic year compared to last year?
- Are incident counts increasing, decreasing, or stable over the past five years?
- Which months have the highest incident volume? Is this consistent year to year?
- Are trends similar across all residence halls, or concentrated in specific buildings?
- How does Fall semester compare to Spring semester?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Incident counts over time | FactIncident | INCIDENT_DATE, joined to DimDate |
| Counts by location | FactIncident | INCIDENT_LOCATION |
| Counts by case type | FactIncident | TYPE |
| Term-over-term comparison | FactIncident | Joined to DimAcademicTerm |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Volume trends alone don't tell you if things are "better" or "worse." A decrease in incidents could mean fewer problems, fewer reports, or a policy change. Pair volume trends with context: Did occupancy change? Did reporting processes change? Did we add or remove buildings from scope?\

**For Analysts:** When comparing years, ensure you're comparing complete periods. If it's October 2024, comparing "AY2425 so far" to "all of AY2324" will be misleading. Use DimDate to filter to equivalent periods (e.g., Aug 1–Oct 15 of each year).
:::

## Charges and Findings

These questions help you understand *what violations are being alleged* and *what findings result*.

**Example Questions:**

- What are the most common violation types?
- Has the mix of violation types changed over time?
- What is the "not responsible" rate overall? By violation type?
- Are we charging more violations per incident than in prior years?
- Do outcome rates differ by case type (e.g., residential vs. academic)?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Violation frequency | FactViolation | VIOLATION |
| Outcomes by violation | FactViolation | VIOLATION, OUTCOME |
| Violations per incident | FactViolation + FactIncident | Count violations, count incidents, divide |
| Outcomes by case type | FactViolation | TYPE, OUTCOME |

**Understanding Violations Per Incident:**

This metric helps distinguish behavior patterns from charging patterns:

| **Scenario** | **What It Might Mean** |
|----------|--------------------|
| Violations per incident increasing | Staff may be charging more thoroughly, or incidents may be more complex |
| Violations per incident decreasing | Staff may be charging more conservatively, or incidents may be simpler |
| High not-responsible rate for specific violation | Evidence standards may be difficult to meet, or the violation may be over-charged |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Outcome differences can reflect many things: behavior, evidence quality, policy clarity, charging patterns, or hearing officer interpretation. A high not-responsible rate for a specific violation isn't automatically a problem—it might mean staff are appropriately bringing close cases forward. Use outcome patterns as *prompts for conversation*, not conclusions.\

**For Analysts:** The OUTCOME field contains the finding (Responsible, Not Responsible, etc.). Some violations may have NULL outcomes if the case is still pending. Filter to resolved cases when calculating outcome rates: `filter(!is.na(OUTCOME))` or equivalent in DAX.
:::

## Sanctions and Interventions

These questions help you understand *what we do in response* to conduct incidents.

**Example Questions:**

- Which sanctions are most frequently applied?
- How has our sanction mix changed over time?
- How often do we use educational interventions vs. status sanctions (probation, suspension)?
- Do sanction patterns differ by violation type or case type?
- After adding a new educational program, how often is it being assigned?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Sanction frequency | FactSanction | Sanction |
| Sanctions over time | FactSanction | INCIDENT_DATE joined to DimDate |
| Sanction categories | FactSanction + DimSanction | Join on Sanction to get CATEGORY_LABEL, SEVERITY_LEVEL |
| Sanctions by case type | FactSanction | TYPE |

**Understanding Sanction Categories (from DimSanction):**

| **Category** | **Examples** | **Purpose** |
|----------|----------|---------|
| Educational | Online modules, reflection papers, seminars | Learning and behavior change |
| Status | Probation, reprimand | Formal record without restriction |
| Restrictive | Loss of privileges, no-contact orders | Limit access or behavior |
| Removal | Suspension, housing termination | Separation from community |
| Referral | CAPS consultation, academic advisor | Connection to support services |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Sanctions represent your intervention philosophy in action. If you believe in educational approaches but 80% of sanctions are probation, there may be a gap between philosophy and practice—or probation may be serving an educational purpose in your context. Use sanction data to prompt training conversations and resource planning.\

**For Analysts:** FactSanction has one row per sanction applied. A single case can have multiple sanctions (e.g., probation + educational program). To count cases with any educational sanction, you'll need to aggregate: count distinct FILE_ID where Sanction is in the educational category.
:::

## Process and Timeline

These questions help you understand *how efficiently* cases move through the system.

**Example Questions:**

- How long do cases typically take from report to resolution?
- Where do delays occur—at intake (report to case creation) or adjudication (case creation to resolution)?
- Are timelines getting longer or shorter over time?
- Do certain case types take longer than others?
- Are there seasonal patterns (e.g., cases taking longer during busy periods)?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Overall timeline | FactTimeline | reported_to_resolution |
| Intake speed | FactTimeline | reported_to_case |
| Adjudication speed | FactTimeline | case_to_resolution |
| Timelines by case type | FactTimeline + FactIncident | Join on FILE_ID to get TYPE |
| Timelines over time | FactTimeline | INCIDENT_DATE joined to DimDate |

**Understanding the Three Timeline Metrics:**

| **Metric** | **What It Measures** | **Influenced By** |
|--------|------------------|---------------|
| reported_to_case | How quickly we create cases from reports | Intake staffing, triage process, reporting clarity |
| case_to_resolution | How quickly we resolve created cases | Hearing officer availability, case complexity, student scheduling |
| reported_to_resolution | Total time in system | All of the above |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Longer timelines aren't automatically bad. A complex Title IX case *should* take longer than a simple noise violation. What matters is whether timelines are *appropriate* for the case type and *consistent* across similar cases. Look for outliers and investigate: Why did this case take 90 days when similar cases take 14?\

**For Analysts:** Timeline metrics have already been adjusted for institutional pause periods (breaks, holidays) as defined in `config.R`. Negative values have been floored to zero. Warning Letter cases skip timeline adjustment since they resolve immediately. When reporting averages, consider using median instead of mean—timeline distributions are typically right-skewed with some very long cases.
:::

## Housing Context and Rates

These questions help you understand conduct patterns *in relation to where students live* and *how many students live there*.

**Example Questions:**

- Which residence halls have the most violations?
- Which halls have the highest violation *rate* per 100 residents?
- Are violation rates higher in certain building types (traditional vs. suite-style)?
- How have rates changed in a specific building over time?
- If we implemented a prevention program in a hall, did rates change afterward?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Violations per hall | FactViolation | INCIDENT_LOCATION |
| Violation rates | FactHousingCensus | VIOLATIONS, CENSUS, Violation_Rate |
| Building characteristics | DimHousing | Housing_Type, Ownership, Room_Style |
| Year-specific context | DimHousingYear | Attributes that change annually |
| Rate trends over time | FactHousingCensus | Academic_Year, Semester |

**Understanding FactHousingCensus:**

This table pre-calculates violation rates so you don't have to:

| **Field** | **What It Contains** |
|-------|------------------|
| INCIDENT_LOCATION | Building name |
| Academic_Year | AY2324, etc. |
| Semester | FA or SP (Summer excluded due to low census) |
| CENSUS | Number of residents that term |
| VIOLATIONS | Count of violations that term |
| Violation_Rate | (VIOLATIONS ÷ CENSUS) × 100 |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Rates enable fair comparisons. A large hall will almost always have more violations than a small hall—that's just math. Rates tell you whether a building has *disproportionate* activity given its size. However, rates can be unstable in small buildings: a hall with 50 residents and 5 violations (10%) could drop to 2% next year just by chance.\

**For Analysts:** FactHousingCensus only includes Fall and Spring terms because summer census is often too small for meaningful rates. If a building has zero census (closed, not yet open), Violation_Rate will be NULL, not zero—this prevents divide-by-zero errors and avoids implying "no violations" when the building wasn't occupied.
:::

## Equity and Disaggregation

These questions help you understand whether conduct patterns *differ across student populations*.

**Example Questions:**

- Are certain student groups over-represented in conduct incidents relative to their enrollment?
- Do outcome rates (responsible/not responsible) differ by student demographics?
- Are sanction patterns consistent across student groups, or are some groups receiving more severe sanctions?
- Do timeline lengths differ by student demographics?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Student demographics | DimStudent | ETHNICITY, GENDER, CLASSIFICATION |
| Academic context | DimStudent | College, Major |
| Involvement by group | FactIncident + DimStudent | Join on SID |
| Outcomes by group | FactViolation + DimStudent | Join on FILE_ID, then to DimStudent |
| Sanctions by group | FactSanction + DimStudent | Join on FILE_ID, then to DimStudent |

**Important Considerations:**

| **Consideration** | **Why It Matters** |
|---------------|----------------|
| Comparison population | Are you comparing to enrolled population, residential population, or something else? |
| Small numbers | Disaggregated data for small groups may not be reliable and could risk identification |
| Intersectionality | A student's experience may be shaped by multiple identities simultaneously |
| Causation | Disparities are patterns to investigate, not conclusions about cause |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Equity analysis is essential but requires careful interpretation. A disparity is a *finding that demands investigation*, not proof of bias or discrimination. The data shows "what"—understanding "why" requires deeper inquiry into policies, practices, reporting patterns, and context.\

**For Analysts:** DimStudent contains one row per unique student using the *most recent* record. If a student's classification changed (Freshman → Sophomore), DimStudent reflects their current status, not their status at the time of each incident. For true longitudinal demographic analysis at time of incident, you would need to preserve demographics per incident—a potential enhancement for future versions.
:::

## Repeat Involvement (Recidivism)

These questions help you understand *how often students return* to the conduct system. This model provides two complementary approaches to measuring recidivism:

| **Approach** | **Table** | **Question It Answers** |
|----------|-------|---------------------|
| Within-year | FactRecidivism | "How many students had multiple cases this academic year?" |
| Cohort-based | FactCohortRecidivism | "Of students who first got in trouble in a given year, how many ever came back?" |

**Example Questions:**

***Within-Year (FactRecidivism):***

- What percentage of students have multiple responsible findings in the same academic year?
- Is the within-year repeat rate increasing or decreasing over time?

***Cohort-Based (FactCohortRecidivism):***

- Of students whose first responsible finding was in AY2122, what percentage had any subsequent responsible finding?
- How quickly do students recidivate (days between first and second case)?
- Do certain student groups have higher long-term recidivism rates?
- Are certain interventions associated with lower long-term recurrence?
- How intense is repeat involvement (one repeat vs. three+ repeats)?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Within-year repeat rates | FactRecidivism | Academic_Year, Found_Resp, Found_Resp_Again |
| Long-term recidivism by cohort | FactCohortRecidivism | Cohort_Year, Is_Recidivist |
| Time to recidivism | FactCohortRecidivism | Days_to_Recidivism |
| Intensity of repeat involvement | FactCohortRecidivism | Recidivism_Category, Repeat_Count |
| Recidivism by student group | FactCohortRecidivism + DimStudent | Join on SID |

### Understanding FactRecidivism (Within-Year):

| **Field** | **What It Contains** |
|-------|------------------|
| Academic_Year | The academic year being measured |
| Found_Resp | Count of unique students found responsible in at least one case |
| Found_Resp_Again | Count of unique students found responsible in more than one case |
| (Rate) | Found_Resp_Again ÷ Found_Resp (calculated in dashboard) |

**Definition:** A student is a within-year recidivist if they were found responsible in **two or more separate cases** within the **same academic year**. Multiple responsible findings in a single case do not count as recidivism.

### Understanding FactCohortRecidivism (Long-Term):

| **Field** | **What It Contains** |
|-------|------------------|
| SID | Student identifier (anonymized) |
| Cohort_Year | Academic year of student's first responsible finding |
| First_Incident_Date | Date of first responsible finding |
| Total_Responsible_Cases | Total cases with responsible findings (all time) |
| Repeat_Count | Number of cases after the first (0 = no recidivism) |
| Is_Recidivist | TRUE if student had any subsequent responsible finding |
| Recidivism_Category | Intensity: No Recidivism, One Repeat, Two Repeats, Three+ Repeats |
| Second_Incident_Date | Date of second responsible finding (NULL if none) |
| Days_to_Recidivism | Days between first and second case (NULL if none) |

**Definition:** A student is assigned to a **cohort** based on the academic year of their **first-ever responsible finding**. They are a recidivist if they have **any subsequent case** with a responsible finding, regardless of when it occurs.

### Comparing the Two Approaches:

| **Aspect** | **FactRecidivism** | **FactCohortRecidivism** |
|-----------|----------------|--------------------|
| Question answered | "How many repeat within a year?" | "How many ever come back?" |
| Time window | Single academic year | All time after first finding |
| Grain | One row per academic year | One row per student |
| Comparable to | Annual repeat rate | Criminal justice recidivism |
| Best for | Short-term patterns, annual reporting | Long-term trends, intervention evaluation |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Recidivism rates are one signal among many. A student returning to the conduct system isn't automatically a "failure" — it might mean they're still learning, they're in a challenging environment, or they have needs that conduct sanctions alone can't address. Use recidivism data to prompt questions about intervention effectiveness and student support.\

Recidivism reflects continued system contact, not student intent, motivation, or responsiveness to support, and should not be interpreted as individual failure.\

**For Analysts:** FactRecidivism provides aggregate counts by year — useful for dashboards but limited for deeper analysis. FactCohortRecidivism is at the student level, enabling joins to DimStudent for demographic disaggregation and to FactSanction (via First_File_ID) for intervention analysis.\

To analyze "do students who receive sanction X have lower recidivism than sanction Y?", join FactCohortRecidivism to FactSanction on the first case, then compare Is_Recidivist rates by sanction type. Note that this is observational — differences may reflect case severity, student characteristics, or other confounds, not sanction effectiveness.
:::

:::{callout-important}
## *Caution: Cohort Maturity*
Recent cohorts will always appear to have lower recidivism rates than older cohorts — not because interventions are working better, but because less time has passed for those students to return. The AY2425 cohort might show 5% recidivism while AY2122 shows 15%, but that gap largely reflects follow-up time, not a meaningful trend.\
When reporting cohort recidivism, either compare cohorts with similar follow-up windows or explicitly note this limitation alongside the figures.
:::

## Workload and Operational Questions

These questions help you understand *how work is distributed* and *where operational attention is needed*.

**Example Questions:**

- How many cases does each hearing officer handle per semester?
- Are case assignments balanced across staff, or are some officers overloaded?
- How many cases are currently pending? Overdue?
- Which case types consume the most staff time?

**Where to Look:**

| **Question Focus** | **Primary Table** | **Key Fields** |
|----------------|---------------|------------|
| Cases per officer | DimCase | ASSIGNED_TO, ASSIGNED_TO_NAME |
| Officer details | DimHearingOfficer | Department, Position |
| Pending/overdue cases | FactDeadline | DEADLINE, Overdue |
| Case volume by type | FactIncident or DimCase | TYPE |

::: {.callout-note}
## *Interpretation Guidance*

**For Decision-Makers:** Workload metrics support staffing conversations but don't capture everything. A hearing officer with fewer cases might handle more complex cases, supervise student staff, or have other responsibilities. Use workload data as a starting point for conversation, not a performance metric.\

**For Analysts:** FactDeadline includes an `Overdue` flag calculated as `DEADLINE < today()`. This is calculated at the time the pipeline runs, so it reflects the state as of the last data refresh. For real-time monitoring, consider connecting Power BI directly to the source or scheduling frequent refreshes.
:::

\newpage

# Adapting This Pipeline

This pipeline was built on one institution's Maxient configuration, data 
structure, and organizational context. Before running it elsewhere, `config.R` 
requires deliberate review and updates. Some changes are required for the 
pipeline to run at all; others depend on how your institution has configured 
Maxient and structured your data.

::: {.callout-important}
`config.R` is designed to be shared across pipelines. `ISSUE_REPLACEMENTS` belongs to a separate behavioral intervention pipeline and can be ignored or removed if you are only implementing the conduct data model. Most of this data model may be adapted to process and visualize CARE/BIT data.
:::

## Updating config.R

**`ACADEMIC_YEARS`** defines which academic years are included in the 
analysis. Update this vector to reflect the years present in your exports. 
The pipeline uses this list to scope recidivism calculations and filter cohort 
tables — years not listed here will be excluded from those outputs even if 
they appear in your data.

**`DATE_RANGE`** sets the start and end dates for the DimDate calendar table. 
This should span your full analysis period with some buffer on each end — 
typically August 1 of your earliest academic year through July 31 of your 
latest. If DimDate doesn't cover a date in your data, fact table joins to 
DimDate will silently fail for those records.

**`PATHS`** contains all file paths for imports, exports, and the pepper key. 
Every path in this list will need to be updated to reflect your local directory 
structure. The pepper key path is particularly sensitive — see the 
Anonymization section for guidance on where to store it and how to protect it.

**`RED_LOCATIONS`** is the list of residential location names the pipeline 
retains after filtering. Only records with an `INCIDENT_LOCATION` matching a 
value in this list will be included in any downstream table. This list must 
reflect your standardized location names — the names that result *after* 
`LOCATION_MAP` has been applied, not the raw values from Maxient. If a valid 
residential location is missing from this list, all incidents at that location 
will be silently dropped.

**`LOCATION_MAP`** standardizes the many variations of location names that 
Maxient produces over time — inconsistent capitalization, HTML-encoded 
characters, building name changes, and abbreviations. Review your raw Maxient 
exports carefully and build this map to reflect the variations present in your 
data. New exports will introduce new variations; add them here before each 
pipeline run. Unmapped locations that are not in `RED_LOCATIONS` are dropped 
silently, so monitoring for unexpected increases in record loss after adding 
new data is important.

**`OMIT_RANGES`** defines the institutional pause periods subtracted from 
timeline calculations — semester breaks, holidays, and any other periods when 
the conduct process is not actively moving forward. These dates are institution-
and year-specific; replace them entirely with your own academic calendar. 
Missing a break period means timeline metrics will overstate actual processing 
time for cases spanning that period. The comments in this section identify the 
type of each pause period to help you match them to your own calendar.

The extensive list in this pipeline reflects a specific institutional policy requirement: the conduct process cannot move forward when classes are not in session, making every break and holiday a legitimate exclusion rather than a convenience. Review your own institutional policy before building this list — the right list is the one your conduct process actually requires, which may be considerably shorter than what is shown here.

Before running the pipeline on new data for the first time, examine the raw 
timeline distributions before the floor is applied. Negative values that cannot 
be explained by pause period overlap indicate date entry errors — hearing dates 
recorded before incident dates, missing case creation dates, and similar 
problems — that will otherwise be silently absorbed and appear as zero-day cases
in the output.

**`VIOLATION_REPLACEMENTS`** standardizes charge names across years of data. 
Maxient exports can produce the same violation in multiple forms — 
HTML-encoded ampersands, sub-type variations, or names that changed when your 
policy was revised. This map consolidates those variations into consistent 
labels for analysis. Your violation names will differ from those shown here; 
build this map from your own export history, paying particular attention to 
charges that appear inconsistently across years.

## Institution-Specific Pipeline Elements

Several parts of the pipeline reflect this institution's Maxient configuration directly and require changes beyond config.R. This section identifies those elements and explains how to build the lookup tables that support them.

### Sanction and Violation Column Names

The `sanctions` extraction step selects sanction columns by their Maxient 
field names (e.g., `ALDRPAPER`, `CANNABISCRS`, `REDSUSP`). These names are 
specific to how an institution has configured its sanction fields in Maxient. Replace the 
column list with the sanction field names from your own export. The column 
selection block is clearly marked in the pipeline and the pattern is 
straightforward: include every sanction column you want to track, and the 
subsequent `pivot_longer()` step will reshape them into the long-format 
FactSanction table automatically.

### Tag Fields

The `tags` extraction step selects tag columns by name, and those names 
reflect a specific institution's Maxient tag configuration. Replace them with the tag field names 
from your export, or remove the tags section entirely if your institution does 
not use Maxient tags. If you add or remove tags, update FactTag and DimTag 
accordingly.

### Institution-Specific Custom Fields

The `OTHER2`, `OTHER3`, and `OTHER4` field renames reflect institution-specific custom 
fields configured in Maxient (`College_2`, `Regional_Campus`, `Citizenship`). 
Rename these to match whatever your institution stores in those fields, or 
remove the rename step if your institution does not use them.

### Building the Lookup Tables

The four lookup tables below require the most institution-specific work. Each 
has a defined structure that the pipeline expects — the column names and formats 
must match what the pipeline reads — but the content must reflect your 
institution's buildings, sanctions, charges, and staff.

**`DimSanction`** maps Maxient sanction abbreviations to descriptive names and 
adds analytical fields for categorization, severity, and flags used in 
dashboards. To build it, export the **Manage Sanctions** screen in Maxient, 
which provides the sanction abbreviation and descriptive name for every sanction 
configured in your system. From there, add the analytical fields your 
dashboards require — at minimum, a category label (e.g., Educational, Status, 
Restrictive, Removal), a severity level, and any flags relevant to your 
analysis (e.g., `IS_RESTRICTIVE`). The category and severity classifications 
reflect your institution's intervention philosophy and should be defined 
deliberately rather than copied from another institution's schema.

**`DimViolation`** maps charge names to policy areas, behavior categories, 
severity levels, and flags used in dashboards. To build it, export the 
**Manage Charges** screen in Maxient, which provides every charge name 
configured in your system. From there, add analytical fields as needed — 
policy area, behavior category, severity level, and binary flags for specific 
analysis needs (e.g., `IS_ALCOHOL`, `IS_VIOLENCE`). As with DimSanction, 
the classifications here reflect your institution's policy structure and 
should be developed with input from conduct professionals who understand how 
charges are used in practice.

**`hearing_officers`** maps Maxient hearing officer assignment codes to staff 
names and department information. To build it, run **Maxient Analytic 914 — 
Hearing Officer Titles in Use**, which returns the assignment code 
(`ASSIGNED_TO`) and staff name (`ASSIGNED_TO_NAME`) columns for every hearing 
officer title attached to a case in your system. From there, add 
whatever analytical fields your dashboards require — position title, 
department, staff type, or groupings that reflect your organizational 
structure. For example, if you want to analyze workload by role rather than 
individual, you can add a `Role_Group` field that assigns all Hall Directors 
to one category, all Assistant Directors to another, and so on.

**`housing_census`** provides resident counts by building and term and is used 
to calculate violation rates in FactHousingCensus. The expected format is one 
row per building with one column per term (e.g., `FA23`, `SP24`). This data 
typically comes from University Housing or a comparable office. See the 
FactHousingCensus section for additional guidance on the expected input format.

### DimHousing and DimHousingYear

These tables describe your residential buildings and must be built manually 
from your institutional knowledge of the housing portfolio. `DimHousing` 
contains stable building attributes — building name, type (traditional, suite, 
apartment), ownership (university-owned, leased, affiliated), and room style. 
`DimHousingYear` captures attributes that change from year to year, such as 
capacity or population type (first-year only, upperclass, mixed). Maintaining 
these as separate tables preserves historical context — you can see what a 
building was like in a given year, not just what it is now.

\newpage

# Data Transformation

This section contains the actual code that transforms raw Maxient exports into the star schema. Each subsection includes:

- A plain-language explanation of what's happening and why
- The R code that implements it
- A schema output showing the resulting table structure (where applicable)

**If you're not reading the code:** Focus on the explanations and schema outputs. They tell you what each step accomplishes without requiring you to understand the syntax.

**If you are reading the code:** The code is designed to be readable and maintainable. Related operations are grouped together, and comments explain non-obvious choices. Please note that this code is not directly transferable to other institutional contexts. Each institution configures its datafeed to Maxient differently and that will impact how it needs to be cleaned and tidied.

## Setup: Load Libraries and Custom Functions

This section loads the software packages and configuration settings needed to transform raw Maxient exports into reporting-ready tables. The `config.R` file contains all constants (academic years, file paths, location lists, replacement mappings) while `conduct_funcs.R` provides custom functions for data cleaning, anonymization, and analysis.

**What This Does:**

- Clears any previous data from memory (ensures a clean start)
- Loads publicly available R packages for data manipulation, Excel export, and secure hashing
- Loads custom functions written specifically for this pipeline
\
```{r message=FALSE, warning=FALSE}
# ─────────────────────────────────────────────────────────────────────────────
# Start with a clean environment
# ─────────────────────────────────────────────────────────────────────────────
rm(list = ls())

# ─────────────────────────────────────────────────────────────────────────────
# Anonymization Toggle
# ─────────────────────────────────────────────────────────────────────────────

ANONYMIZE <- TRUE

# ─────────────────────────────────────────────────────────────────────────────
# Publicly Available Packages
# ─────────────────────────────────────────────────────────────────────────────
library(tidyverse)
library(writexl)
library(openssl)
library(digest)

# ─────────────────────────────────────────────────────────────────────────────
# Custom Functions and Configuration (included in appendices)
# ─────────────────────────────────────────────────────────────────────────────
source("~/GitHub/conduct-analytics/shared/R/conduct_funcs.R")
source("~/GitHub/conduct-analytics/shared/R/config.R")
```

::: {.callout-tip}
## *Why This Matters*

**For Decision-Makers:** This setup ensures that every time the pipeline runs, it starts fresh and uses the same tools. That's what makes results reproducible—you'll get the same output whether you run it today or next month.\

**For Analysts:** The custom functions in `conduct_funcs.R` handle common operations (date conversion, anonymization, schema display). The configuration in `config.R` contains all mappings, date ranges, and constants. Separating these from the main pipeline makes maintenance easier—you can update a location mapping without touching the transformation logic.
:::

## Data Import

This section brings in all the raw data: multiple Maxient exports and the lookup tables that add institutional context.

### Import Maxient Exports

Maxient exports are limited in size, so historical data spans multiple files. The pipeline reads each file and combines them into one dataset.
\
```{r message=FALSE, warning=FALSE}
# ─────────────────────────────────────────────────────────────────────────────
# Load Raw Maxient Exports
# ─────────────────────────────────────────────────────────────────────────────

df1 <- read_csv(
  PATHS$imports$import1,
  show_col_types = FALSE
)

df2 <- read_csv(
  PATHS$imports$import2,
  show_col_types = FALSE
)

df3 <- read_csv(
  PATHS$imports$import3,
  show_col_types = FALSE,
  col_types = cols(
    .default = col_guess(),
    OTHER2 = col_character(),
    OTHER3 = col_character(),
    OTHER4 = col_character(),
    `Tag: Residence Halls Vandalism` = col_character(), 
    `Tag: Hospital Transport` = col_character(), 
    `Tag: Health & Safety Inspection` = col_character(), 
    `Tag: Report Dismissed/Complainant Form Not Received` = col_character(), 
    `Tag: Case Dismissed by SCCS Determination` = col_character(), 
    `Tag: Case Dismissed/Complainant Non-Compliant` = col_character()
  )
)

# Combine all exports into one dataset
df <- bind_rows(df1, df2, df3)

# Free up memory by removing individual files
rm(df1, df2, df3)
```

### Import Lookup Tables

Lookup tables add institutional context that isn't stored in Maxient directly. Depending on how an institution has configured Maxient, some fields may pass through codes rather than human-readable values. In this pipeline, academic majors are recorded as codes rather than names. The academic plans lookup translates that code into three meaningful pieces of information: the academic organization code (which college the student belongs to), the academic career (undergraduate or graduate), and the academic plan name (the actual major). A separate college lookup then maps organization codes to college names. A third lookup translates hearing officer assignment codes into staff names and department information.

It's worth noting that updating a Maxient configuration to pass through richer data only affects cases created after that change. Cases created before the update retain whatever was recorded at the time. For institutions with years of historical data, external lookup tables are often the only reliable way to decode legacy records consistently across the full analysis period.

Together these tables make it possible to answer questions like "Which colleges have the highest involvement rates?" and "How is caseload distributed across staff?" that the raw Maxient export alone cannot support.
\
```{r message=FALSE, warning=FALSE}
# ─────────────────────────────────────────────────────────────────────────────
# Lookup Tables
# ─────────────────────────────────────────────────────────────────────────────

# Academic Plan Codes → College names and major descriptions
DimAcademicPlans <- read_csv(PATHS$imports$academic_plans, show_col_types = FALSE,
                col_types = cols(
                  `Academic Organization Code` = col_character()
                ))
DimCollege <- read_csv(PATHS$imports$dim_college, 
                       show_col_types = FALSE,
                       col_select = -NOTES,
                       col_types = cols(
                         `Academic Organization Code` = col_character()
                       ))

# Hearing Officer assignment codes → Staff names, departments, positions
ho <- read_csv(PATHS$imports$hearing_officers, show_col_types = FALSE)
```

::: {.callout-tip}
## *Why Lookup Tables?*

**For Decision-Makers:** Maxient stores major as a code (like "15BSBA"). That's not helpful for analysis. The lookup table translates it to "Business Administration" and tells us it's in the Lindner College of Business. This lets dashboards answer questions like "Which colleges have the highest involvement rates?"\

**For Analysts:** Lookup tables are joined using `left_join()`. Unmatched values become NA, which we then replace with "Not Reported." If you see unexpected "Not Reported" values increasing, check whether the lookup tables need updating (new majors, new staff, etc.).
:::

## Tidy and Enrich Core Data

This section transforms the raw combined dataset into something usable for analysis. Each subsection addresses a specific data quality issue.

### Standardize Locations and Join Context

**What This Does:**

- Converts location names to title case for consistency
- Applies the location mapping to standardize variations (e.g., "CALHOUN" → "Calhoun Hall")
- Filters to only residential locations (defined in `config.R`)
- Converts date columns from text to proper date format
- Joins lookup tables to add academic plan, college, and hearing officer context
- Replaces missing values with explicit "Not Reported" or "Unheard" labels
\
```{r message=FALSE, warning=FALSE}
df <- df |>
# ───────────────────────────────────────────────────────────────────────────
# Standardize location names
# ───────────────────────────────────────────────────────────────────────────
mutate(
    INCIDENT_LOCATION = fix_nbsp(INCIDENT_LOCATION),
    INCIDENT_LOCATION = str_to_title(INCIDENT_LOCATION),
    INCIDENT_LOCATION = recode(INCIDENT_LOCATION, !!!LOCATION_MAP,
                               .default = INCIDENT_LOCATION)
  ) |>
  # Keep only residential locations
  filter(INCIDENT_LOCATION %in% RED_LOCATIONS) |>
  
  # ───────────────────────────────────────────────────────────────────────────
  # Convert date columns
  # ───────────────────────────────────────────────────────────────────────────
  convert_to_date(c(
    "APPT_DATE", "CASE_CREATED_DATE", "DOB", "REPORTED_DATE",
    "INCIDENT_DATE", "HEARING_DATE", "DEADLINE"
  )) |>
  
  # ───────────────────────────────────────────────────────────────────────────
  # Join lookup tables for institutional context
  # ───────────────────────────────────────────────────────────────────────────
  left_join(DimAcademicPlans, by = c("ACADEMIC_MAJOR" = "Academic Plan Code")) |>
  left_join(ho, by = "ASSIGNED_TO") |>
  left_join(DimCollege, by = "Academic Organization Code") |>

  # ───────────────────────────────────────────────────────────────────────────
  # Rename columns for clarity
  # ───────────────────────────────────────────────────────────────────────────
  rename(
    Resolution         = HEARING_TYPE,
    ACADEMIC_PLAN      = `Academic Plan`,
    ACADEMIC_PLAN_CODE = ACADEMIC_MAJOR,
    ACADEMIC_ORG_CODE  = `Academic Organization Code`
  ) |>

  # ───────────────────────────────────────────────────────────────────────────
  # Replace NA values with explicit labels
  # ───────────────────────────────────────────────────────────────────────────
  mutate(
    Resolution         = replace_na(Resolution, "Unheard"),
    College            = replace_na(College, "Not Reported"),
    ACADEMIC_PLAN      = replace_na(ACADEMIC_PLAN, "Not Reported"),
    ACADEMIC_PLAN_CODE = replace_na(ACADEMIC_PLAN_CODE, "Not Reported"),
    ACADEMIC_ORG_CODE  = replace_na(ACADEMIC_ORG_CODE, "Not Reported"),
    ASSIGNED_TO_NAME   = replace_na(ASSIGNED_TO_NAME, "SCCS Staff"),
    CLASSIFICATION     = replace_na(CLASSIFICATION, "Not Reported")
  )
```

::: {.callout-tip}
## *Why Explicit Missing Values?*

**For Decision-Makers:** When a filter dropdown shows "Not Reported," you know there's data with missing information—and you can decide whether to include it. If we left these as blank, they'd silently disappear from many visualizations, potentially skewing results.\

**For Analysts:** The `replace_na()` calls happen *after* the joins, catching both originally-missing values and values that failed to match in lookup tables. If "Not Reported" counts increase unexpectedly, investigate whether lookup tables are out of date.
:::

### Standardize Categorical Values

**What This Does:**

- Recodes classification values for consistency (e.g., "Pre-Junior" → "Sophomore")
- Standardizes athletics sport names (various abbreviations → consistent format)
- Standardizes role labels (e.g., "Alleged/Respondent" → "Respondent")
\
```{r}
df <- df |>
  mutate(
    # ─────────────────────────────────────────────────────────────────────────
    # Standardize classification
    # ─────────────────────────────────────────────────────────────────────────
    CLASSIFICATION = recode(CLASSIFICATION, 
                            "Pre-Junior" = "Sophomore",
                            "First Year (1L)" = "Graduate",
                            "Second Year (2L)" = "Graduate",
                            "Third Year (3L)" = "Graduate",
                            "Student" = "Not Reported"),
    
    # ─────────────────────────────────────────────────────────────────────────
    # Standardize athletics sport names
    # ─────────────────────────────────────────────────────────────────────────
    MEM_ATHLETICS_SPORT = recode(
      MEM_ATHLETICS_SPORT,
      "Men's Track &amp; Field"          = "Men's Track and Field",
      "Women's Track &amp; Field"        = "Women's Track and Field",
      "Women's Indoor Track &amp; Field" = "Women's Track and Field",
      "M-Football"                        = "Men's Football",
      "M-Baseball"                        = "Men's Baseball",
      "Baseball"                          = "Men's Baseball",
      "Cheerleaders"                      = "Cheer",
      "Football"                          = "Men's Football",
      "Manager"                           = "Athletic Manager",
      "Men's Swimming &amp; Diving"      = "Men's Swim and Dive",
      "N/A"                               = "Not Athlete",
      "Women's Swimming &amp; Diving"    = "Women's Swim and Dive",
      "Trainer"                           = "Athletic Trainer"
    ),
    
    # ─────────────────────────────────────────────────────────────────────────
    # Standardize role labels
    # ─────────────────────────────────────────────────────────────────────────
    ROLE = recode(ROLE, "Alleged/Respondent" = "Respondent")
  ) |>
  filter(CLASSIFICATION %nin% c('Staff', 'Student Organization'))
```

### Calculate Timeline Metrics

**What This Does:**

Calculates three measures of how long cases take to move through the process:

| **Metric** | **Measures** |
|--------|----------|
| `reported_to_case` | Days from incident report to case creation |
| `case_to_resolution` | Days from case creation to hearing/resolution date |
| `reported_to_resolution` | Total days from report to resolution |

The calculations subtract institutional pause periods (breaks, holidays) and prevent negative values from data entry errors.

Institutions seeking to replicate this approach should verify that these date fields map correctly to their own process. The sequence assumed here is: incident reported → case created → case resolved. If your workflow uses different date fields, records them in a different order, or defines "resolution" differently, the metric definitions will need to be adjusted accordingly before the results are meaningful.
\
```{r}
df <- df |>
  mutate(
    # ─────────────────────────────────────────────────────────────────────────
    # Calculate raw day differences, then subtract pause periods
    # ─────────────────────────────────────────────────────────────────────────
    reported_to_case = 
      as.numeric(difftime(CASE_CREATED_DATE, REPORTED_DATE, 
                          units = "days")) -
      compute_overlaps(REPORTED_DATE, CASE_CREATED_DATE, Resolution, 
                       OMIT_RANGES),
    
    case_to_resolution = 
      as.numeric(difftime(HEARING_DATE, CASE_CREATED_DATE, 
                          units = "days")) -
      compute_overlaps(CASE_CREATED_DATE, HEARING_DATE, Resolution, 
                       OMIT_RANGES),
    
    reported_to_resolution = 
      as.numeric(difftime(HEARING_DATE, REPORTED_DATE, 
                          units = "days")) -
      compute_overlaps(REPORTED_DATE, HEARING_DATE, Resolution, 
                       OMIT_RANGES)
  ) |>
  
  # ───────────────────────────────────────────────────────────────────────────
  # Floor negative values to zero (handles data entry errors)
  # ───────────────────────────────────────────────────────────────────────────
  mutate(
    across(
      c(reported_to_case, case_to_resolution, reported_to_resolution),
      ~ pmax(.x, 0)
    )
  )
```

::: {.callout-tip}
## *Why Adjust for Pause Periods?*

**For Decision-Makers:** Without this adjustment, a case reported December 15 and resolved January 15 would show as 31 days—even though the office was closed for 20 of those days. Subtracting pause periods gives a more accurate picture of *working time* in the system.\

**For Analysts:** Pause periods are defined in `config.R` as `OMIT_RANGES`. The `compute_overlaps()` function (from `conduct_funcs.R`) calculates how many days of each case's timeline fall within these ranges. Warning Letter cases skip adjustment because they resolve immediately. If you add new pause periods, add them to `OMIT_RANGES`.
The floor to zero is a necessary consequence of the pause period adjustment, not just a data quality safeguard. Cases opened and resolved within a pause window — or where the pause period exceeds the raw elapsed time — will produce negative values after adjustment. Flooring those to zero is the correct treatment: the case consumed no active process time by definition.
:::

### Filter Population and Add Academic Time Fields

**What This Does:**

- Removes non-student records (locations, non-students, "Other" roles)
- Creates academic year based on incident date
- Adds time-based fields for trend analysis (week, weekday, month)
- Calculates age at incident
- Renames institution-specific "OTHER" fields to meaningful names
- Standardizes college names to abbreviations
\
```{r}
df <- df |>
  # ───────────────────────────────────────────────────────────────────────────
  # Filter to student respondents only
  # ───────────────────────────────────────────────────────────────────────────
  filter(!ROLE %in% c("Location", "Non-Student", "Non-student", 
                      "Other")) |>
  filter(CLASSIFICATION != "Non-Student") |>
  
  # ───────────────────────────────────────────────────────────────────────────
  # Create academic year and time-based fields
  # ───────────────────────────────────────────────────────────────────────────
  mutate(
    ACADEMIC_YEAR = case_when(
      INCIDENT_DATE >= ymd("2017-08-01") & 
        INCIDENT_DATE <= ymd("2018-07-31") ~ "AY1718",
      INCIDENT_DATE >= ymd("2018-08-01") & 
        INCIDENT_DATE <= ymd("2019-07-31") ~ "AY1819",
      INCIDENT_DATE >= ymd("2019-08-01") & 
        INCIDENT_DATE <= ymd("2020-07-31") ~ "AY1920",
      INCIDENT_DATE >= ymd("2020-08-01") & 
        INCIDENT_DATE <= ymd("2021-07-31") ~ "AY2021",
      INCIDENT_DATE >= ymd("2021-08-01") & 
        INCIDENT_DATE <= ymd("2022-07-31") ~ "AY2122",
      INCIDENT_DATE >= ymd("2022-08-01") & 
        INCIDENT_DATE <= ymd("2023-07-31") ~ "AY2223",
      INCIDENT_DATE >= ymd("2023-08-01") & 
        INCIDENT_DATE <= ymd("2024-07-31") ~ "AY2324",
      INCIDENT_DATE >= ymd("2024-08-01") & 
        INCIDENT_DATE <= ymd("2025-07-31") ~ "AY2425",
      TRUE ~ "AY2526"
    ),
    
    # Week number for week-over-week analysis
    WEEK_NUM = isoweek(INCIDENT_DATE),
    
    # Day of week for day patterns
    incident_wday = wday(INCIDENT_DATE, label = TRUE),
    
    # Month ordered by academic year (Aug–Jul) rather than calendar year
    incident_month = factor(
      month(INCIDENT_DATE, label = TRUE, abbr = TRUE),
      levels = c("Aug", "Sep", "Oct", "Nov", "Dec", 
                 "Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul")
    ),
    
    # Age at time of incident
    age_at_incident = round((DOB %--% INCIDENT_DATE) / years(1), 2)
  ) |>
  
  # ───────────────────────────────────────────────────────────────────────────
  # Rename institution-specific fields
  # ───────────────────────────────────────────────────────────────────────────
  rename(
    College_2       = OTHER2,
    Regional_Campus = OTHER3,
    Citizenship     = OTHER4
  )
```

::: {.callout-tip}
## *Why Filter to Students Only?*

**For Decision-Makers:** The conduct dashboards are designed to analyze *student* conduct. Including "Location" records (used for tracking where something happened without a person attached) or non-student records would inflate counts and create confusing results.\

**For Analysts:** The explicit `case_when()` is used rather than `compute_academic_year()` to ensure only records within the intended analysis period are assigned named academic years. The default case `(TRUE ~ 'AY2526')` captures any records outside the expected range rather than silently extending the analysis. If you extend the data range, add new conditions or update the default case.
:::

### Normalize Ethnicity Values

**What This Does:** Standardizes the many variations of ethnicity codes in the source data into consistent reporting categories.
\
```{r}
df <- df |>
  mutate(
    # ─────────────────────────────────────────────────────────────────────────
    # Preserve raw value, then normalize
    # ─────────────────────────────────────────────────────────────────────────
    ETHNICITY_raw = ETHNICITY,
    ETHNICITY = ETHNICITY |>
      str_trim() |>
      str_replace_all("[,/]+", " ") |>
      str_squish() |>
      toupper()
  ) |>
  mutate(
    # ─────────────────────────────────────────────────────────────────────────
    # Map to standard categories
    # ─────────────────────────────────────────────────────────────────────────
    ETHNICITY_BUCKET = case_when(
      ETHNICITY %in% c("HISPA") ~ "Hispanic",
      ETHNICITY %in% c("AMIND") ~ "American Indian",
      ETHNICITY %in% c("NHOPO") ~ "Native Hawaiian or Pacific Islander",
      ETHNICITY %in% c("BLACK", "BLACK NOTHISP", "AFRICAN-AMERICAN")
      ~ "Black",
      ETHNICITY %in% c("WHITE", "NOTHISP WHITE", "CAUCASIAN") ~ "White",
      ETHNICITY %in% c("ASIAN", "ASIAN NOTHISP") ~ "Asian",
      ETHNICITY %in% c("NSPEC") ~ "Not Reported",
      ETHNICITY %in% c("NOTHISP", "OTHER", "OTHER/NO RESPONSE") ~ "Other",
      
      # Multi-word codes (after normalization) indicate multi-ethnic
      str_detect(ETHNICITY, "\\b[A-Z]+\\b\\s+\\b[A-Z]+\\b") 
      ~ "Multi-Ethnic",
      
      # Fallback cases
      is.na(ETHNICITY) ~ "Not Reported",
      TRUE ~ "Other"
    ),
    
    # ─────────────────────────────────────────────────────────────────────────
    # Convert to ordered factor
    # ─────────────────────────────────────────────────────────────────────────
    ETHNICITY = factor(
      ETHNICITY_BUCKET,
      levels = c("Hispanic", "American Indian", 
                 "Native Hawaiian or Pacific Islander",
                 "Black", "White", "Asian", "Multi-Ethnic", 
                 "Other", "Not Reported")
    )
  ) |>
  select(-ETHNICITY_BUCKET)
```

::: {.callout-tip}
## *Why Standardize Ethnicity?*

**For Decision-Makers:** Equity analysis requires consistent categories. If "Black" students appear as "BLACK," "Black NotHisp," and "African-American" in different records, they'll be split across three categories in any chart—hiding the actual pattern.\

**For Analysts:** The raw value is preserved in `ETHNICITY_raw` for troubleshooting. If new codes appear in future exports, they'll fall into "Other"—monitor for unexpected growth in that category. The factor levels control sort order in visualizations.
:::

### Anonymize Identifiable Fields

**What This Does:** Replaces real identifiers (case numbers, file IDs, student IDs) with scrambled codes that cannot be reversed to reveal the original values.
\
```{r}
#| eval: true

# ─────────────────────────────────────────────────────────────────────────────
# Load the encryption key
# ─────────────────────────────────────────────────────────────────────────────
pepper <- readBin(PATHS$pepper, "raw", 32)

# ─────────────────────────────────────────────────────────────────────────────
# Apply hashing to identifier columns
# ─────────────────────────────────────────────────────────────────────────────
if (ANONYMIZE) {
  df <- df |>
    mutate(
      CASE_NUMBER = hash_col(CASE_NUMBER, pepper),
      FILE_ID     = hash_col(FILE_ID, pepper),
      SID         = hash_col(SID, pepper)
    )
}
```

::: {.callout-tip}
## *How Anonymization Works*

**For Decision-Makers:** After this step, a student ID like "M12345678" becomes something like "a3f7c9e2b1d8...". The scrambled code is consistent—the same student always gets the same code—so you can still track repeat involvement and link records across tables. But no one can work backward from the code to find the real ID.\

**For Analysts:** The `pepper.bin` file is the encryption key. It must be:

- **Protected**: Don't commit to version control or share via email
- **Backed up**: If lost, you can't regenerate consistent hashes
- **Consistent**: Use the same key for every pipeline run, or hashes won't match across refreshes

The `hash_col()` function in `conduct_funcs.R` uses HMAC-SHA256 and returns the first 32 hex characters.
:::

### Prepare Sanctions and Tags for Extraction

**What This Does:** Creates a separate dataset of sanctions that will be used later to build the FactSanction table.
\
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Extract sanction columns for later use
# ─────────────────────────────────────────────────────────────────────────────
sanctions <- df |>
    select(
      FILE_ID, TYPE, INCIDENT_DATE, College, ACAD_ADV_MTG, ADDENDUM, 
      ALDRPAPER, ANGERMGMT, APOLOGY, ASEP, BASICS, BEHAVIORMOD, BULLETIN,
      BULLYING, CANNABISCRS, COMMUNITY_SERVICE, CONFRESCRS, DMS, 
      DRUGALCAWARE, EDU_SIGN, EXPULSION, FIRESAFEED, INTERIMSUSP, 
      LIFESKILLS, LOSSPRIV, NOCONTACT, PARENTNOTIFY, PFLDRUG, REDEXPUL, 
      REDSUSP, REDTRANSFER, REENTRY_MTG, REFLECTCONT, REFLGTUL, REPRIMAND, 
      RESTITUTION, RESTRICTION, SOBEREXP, THEFTAWARE, TOBACCO, VAPING, DP, 
      SUSPENSION)

tags <- df |>
  select(
    FILE_ID,
    `Tag: Case Dismissed by SCCS Determination`,
    `Tag: Case Dismissed/Complainant Non-Compliant`,
    `Tag: Health & Safety Inspection`,
    `Tag: Hospital Transport`,
    `Tag: Report Dismissed/Complainant Form Not Received`,
    `Tag: Residence Halls Vandalism`
  )
```

### Review: Structure After Cleaning

This diagnostic output shows the structure of the cleaned dataset before it's split into dimension and fact tables.

**Reading the output:**

- Lines 1–37: Standard Maxient fields
- Lines 38–42 and 76: institution-specific tag fields
- Lines 43–72 and 77-85: Sanction columns (institution-specific)
- Lines 73–75: Institution-specific "OTHER" fields (renamed above)
- Lines 86+: Fields added through lookups and calculations
\
```{r}
df_schema(df) |>
  select(column, type) |>
  print(n = Inf)
```

\newpage

## Star Schema Tables

This section constructs the final tables that Power BI will use. Each table has a specific purpose and a clear grain (what one row represents).

**Quick Reference:**

| **Table Type** | **Purpose** | **Examples** |
|------------|---------|----------|
| Dimension | Descriptive context for filtering | DimDate, DimStudent, DimHousing |
| Fact | Events and measures to count/analyze | FactIncident, FactViolation, FactSanction |

### Dimensions

Dimension tables store descriptive context that's reused across multiple fact tables. They enable consistent filtering and prevent repeating the same information in every row of the fact tables.

#### DimAcademicPlans

**What It Contains:** A controlled list of academic plan codes with their corresponding academic organization, career level, and major name. Used to translate coded major fields in Maxient exports into human-readable values for analysis.

**Use This When:** You need to filter or group by academic major, college, or career level (undergraduate vs. graduate).
\
```{r}
DimAcademicPlans <- DimAcademicPlans |>
  rename(
    ACADEMIC_ORG_CODE    = `Academic Organization Code`, 
    ACADEMIC_CAREER_CODE = `Academic Career Code`,
    ACADEMIC_PLAN_CODE   = `Academic Plan Code`,
    ACADEMIC_PLAN        = `Academic Plan`
    )
```

#### DimCase

**What It Contains:** One row per FILE_ID, with stable case-level attributes.

**Use This When:** You need to filter or analyze by case characteristics (type, status, assignment) rather than individual incidents or violations.

| **Key Fields** | **Description** |
|------------|-------------|
| FILE_ID | Unique case identifier (anonymized) |
| CASE_NUMBER | Case number (anonymized) |
| TYPE | Case type (Residential, Academic, etc.) |
| STATUS | Current status (Closed, Open, etc.) |
| Resolution | How the case was resolved |
| ASSIGNED_TO | Hearing officer assignment code |
| ASSIGNED_TO_NAME | Hearing officer name |
\
```{r}
DimCase <- df |>
  select(
    FILE_ID,
    CASE_NUMBER,
    TYPE,
    STATUS,
    HOLD_IN_PLACE,
    ASSIGNED_TO,
    ASSIGNED_TO_NAME,
    Resolution,
    CASE_CREATED_DATE
  ) |>
  mutate(
    HOLD_IN_PLACE = case_when(
      HOLD_IN_PLACE == "Yes" ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  distinct()
```

::: {.callout-tip}
## *When to Use DimCase*

**For Decision-Makers:** Use case-level analysis when your question is about workload, staffing, or case characteristics. "How many cases did Officer Smith handle?" is a case-level question.\

**For Analysts:** `distinct()` ensures one row per case. If a case has multiple incidents or students, it still appears once here. Join to fact tables on FILE_ID.
:::

#### DimCollege

**What It Contains:** A controlled list of academic colleges referenced in the data.
\
```{r}
DimCollege <- DimCollege |>
  rename(ACADEMIC_ORG_CODE = `Academic Organization Code`)
```

#### DimDate

**What It Contains:** One row per calendar date, spanning the full analysis period. Includes both calendar fields (year, month, day) and academic time fields (academic year, semester, term).

**Why This Matters:** This is the backbone of all time-based analysis. Every fact table connects to DimDate through a date field, enabling consistent filtering by semester, academic year, or any other time period.

| **Key Fields** | **Description** |
|------------|-------------|
| Date | The calendar date |
| Academic Year | AY format (e.g., "AY2324") |
| Semester | FA, SP, or SU |
| AcademicTermKey | Numeric key for sorting terms correctly |
| AY_Order | Numeric key for sorting academic years |
\
```{r}
DimDate <- build_dim_date(DATE_RANGE$start, DATE_RANGE$end)
```

::: {.callout-tip}
## *Why a Separate Date Table?*

**For Decision-Makers:** A dedicated date table ensures that "Fall 2023" means exactly the same thing on every page of every dashboard. Without it, different visuals might calculate time periods differently, leading to confusion.\

**For Analysts:** In Power BI, mark this as the date table and create relationships from fact tables' date columns to DimDate[Date]. The AcademicTermKey and AY_Order fields enable correct sorting—without them, "AY2324" would sort before "AY1920" alphabetically.
:::

#### DimAcademicTerm

**What It Contains:** One row per academic term (Fall, Spring, Summer for each academic year). Provides labels and sort keys for term-level analysis.
\
```{r}
DimAcademicTerm <- DimDate |>
  filter(Semester %in% c("FA", "SP", "SU")) |>
  distinct(
    Academic_Year = `Academic Year`,
    Semester,
    AcademicTermKey
  ) |>
  mutate(
    AcademicTermLabel = paste(Academic_Year, Semester),
    TermSort          = AcademicTermKey
  ) |>
  arrange(TermSort)
```

#### DimAcademicYear

**What It Contains:** One row per academic year with explicit start/end dates and sort order.
\
```{r}
DimAcademicYear <- DimDate |>
  distinct(`Academic Year`) |>
  rename(ACADEMIC_YEAR = `Academic Year`) |>
  mutate(
    AY_Start_Year = as.integer(paste0("20", str_sub(ACADEMIC_YEAR, 3, 4))),
    AY_Start_Date = as.Date(paste0(AY_Start_Year, "-08-01")),
    AY_End_Date   = as.Date(paste0(AY_Start_Year + 1, "-07-31")),
    AY_Order      = as.integer(str_sub(ACADEMIC_YEAR, 3, 6))
  ) |>
  select(ACADEMIC_YEAR, AY_Order, AY_Start_Date, AY_End_Date) |>
  arrange(AY_Order)
```

::: {.callout-tip}
## *Why Explicit Year Boundaries?*

**For Decision-Makers:** When someone asks "What happened in AY2324?", there's no ambiguity: August 1, 2023 through July 31, 2024. These explicit boundaries prevent the "off by a month" errors that plague calendar-year reporting.\

**For Analysts:** Use AY_Order for sort order in visuals. The start/end dates can be used for date range filtering or for calculating whether a date falls within a specific academic year.
:::

#### DimHearingOfficer

**What It Contains:** Staff metadata for workload and assignment analysis.
\
```{r}
DimHearingOfficer <- ho |>
  distinct()
```

#### DimHousing

**What It Contains:** One row per residence hall with stable building attributes (type, ownership, room style). This data is maintained in partnership with University Housing.
\
```{r message=FALSE}
DimHousing <- read_csv(
  PATHS$imports$dim_housing,
  locale = locale(encoding = "Windows-1252")
) |>
  mutate(INCIDENT_LOCATION = fix_nbsp(INCIDENT_LOCATION)) |>
  arrange(INCIDENT_LOCATION)
```

#### DimHousingYear

**What It Contains:** Housing attributes that can change from year to year (capacity, population type, etc.).

**Why Separate from DimHousing?** A building's capacity or use might change. Separating year-specific attributes preserves historical context—you can see what a building was like in AY2122, not just what it's like now.
\
```{r message=FALSE}
DimHousingYear <- read_csv(
  PATHS$imports$dim_housing_year,
  locale = locale(encoding = "Windows-1252")
) |>
  mutate(INCIDENT_LOCATION = fix_nbsp(INCIDENT_LOCATION)) |>
  arrange(INCIDENT_LOCATION)
```

#### DimSanction

**What It Contains:** Sanction metadata including categories, severity levels, and flags for analysis.

| **Key Fields** | **Description** |
|------------|-------------|
| Sanction | The sanction code |
| CATEGORY_LABEL | Purpose category (Educational, Status, Restrictive, etc.) |
| RESTRICTIVE_FLAG | Whether the sanction restricts student behavior/access |
| SEVERITY_LEVEL | Numeric severity (0=Educational to 4=Removal) |
| SEVERITY_LABEL | Text label for severity level |
\
**Severity Scale:**

| **Level** | **Label** | **Examples** |
|-------|-------|----------|
| 0 | Educational | Online modules, reflection papers, etc. |
| 1 | Status (Non-Restrictive) | Probation, reprimand |
| 2 | Conditional Restriction | Probation, deferred suspension |
| 3 | Restrictive | No-contact orders, loss of privileges |
| 4 | Removal | Suspension, housing termination |
\
```{r message=FALSE}
DimSanction <- read_csv(PATHS$imports$dim_sanction)
```

::: {.callout-tip}
## *Why Categorize Sanctions?*

**For Decision-Makers:** Knowing that we issued 500 sanctions isn't as useful as knowing we issued 350 educational interventions, 100 status sanctions, and 50 restrictive sanctions. Categories help you understand your intervention philosophy in practice.\

**For Analysts:** Join FactSanction to DimSanction on the Sanction field. You can then group by CATEGORY_LABEL or SEVERITY_LEVEL for aggregate analysis. The RESTRICTIVE_FLAG enables quick filtering to "sanctions that limit student access."
:::

#### DimStudent

**What It Contains:** One row per unique student using their most recent record. Includes demographic and academic context for equity analysis. Because this uses the most recent snapshot, a student's attributes (like classification) reflect their current status, not their status at the time of each incident.
\
```{r}
DimStudent <- df |>
  filter(!is.na(SID)) |>
  arrange(SID, INCIDENT_DATE) |>
  group_by(SID) |>
  slice_tail(n = 1) |>
  ungroup() |>
  select(
    SID,
    DOB,
    GENDER,
    ETHNICITY,
    CLASSIFICATION,
    MEM_ATHLETICS_SPORT,
    GPA_CUME,
    ACADEMIC_ORG_CODE,
    ACADEMIC_PLAN_CODE,
    ACADEMIC_PLAN,
    Regional_Campus,
    Citizenship
  )
```

::: {.callout-tip}
## *Privacy and Purpose*

**For Decision-Makers:** This table supports pattern analysis (e.g., "Are any student groups over-represented?"), not individual case lookup. Student IDs are anonymized, so you can see patterns without identifying specific students.\

**For Analysts:** The `slice_tail(n = 1)` after sorting by date ensures we keep the most recent record per student. If you need demographics *at time of incident* for longitudinal demographic analysis, you'd need to preserve that in the fact tables—a potential future enhancement.
:::

#### DimViolation

**What It Contains:**: Charge metadata including categories, severity levels, and flags for analysis.

| **Field** | **Description** |
|-------|-------------|
| VIOLATION | The violation/charge name |
| POLICY | Policy area (Academic Integrity, Alcohol, Drugs, Harassment, Residential, General Conduct, etc.) |
| CATEGORY | Behavior category (Substance Use, Interpersonal Harm, Property, Safety, Administrative, etc.) |
| IS_ALCOHOL | Whether the violation involves alcohol |
| IS_DRUG | Whether the violation involves drugs/narcotics |
| IS_VIOLENCE | Whether the violation involves violence or threat of harm |
| SEVERITY_LEVEL | Numeric severity (1–5) |
| SEVERITY_LABEL | Text label for severity level |
| RATIONALE | Brief explanation of severity classification |

**Severity Scale:**

| **Level** | **Label** | **Examples** |
|-------|-------|----------|
| 1 | Minor | Quiet hours, guest policy, tobacco, passive participation |
| 2 | Moderate | Underage alcohol, paraphernalia, failure to comply, dishonesty |
| 3 | Serious | Cheating, drug possession, theft, fire safety, destruction of property |
| 4 | Severe | Hazing, harassment, weapons, public endangerment, retaliation |
| 5 | Critical | Distribution, sexual misconduct, stalking, physical violence, domestic violence |
\

```{r message=FALSE}
DimViolation <- read_csv(
  PATHS$imports$dim_charge)
```

### Fact Tables

Fact tables store events, outcomes, and measures. Each table has a clear grain which makes counts and calculations reliable.

#### FactDeadline

**Grain:** One row per FILE_ID with a deadline set.

**What It Contains:** Case deadlines and whether they're overdue. Supports operational monitoring.
\
```{r}
FactDeadline <- df |>
  filter(!is.na(DEADLINE)) |>
  mutate(
    Overdue = DEADLINE < today()
  ) |>
  select(
    FILE_ID,
    TYPE,
    STATUS,
    DEADLINE,
    DEADLINE_REASON,
    HOLD_IN_PLACE,
    ASSIGNED_TO,
    ACADEMIC_YEAR,
    Overdue
  )
```

::: {.callout-tip}
## *Real-Time Monitoring*

**For Decision-Makers:** This table answers "What cases need attention right now?" Filter to Overdue = TRUE to see cases past their deadline.\

**For Analysts:** The `Overdue` flag is calculated when the pipeline runs (`today()`). For real-time accuracy, either refresh frequently or calculate overdue status in Power BI using DAX.
:::

#### FactIncident

**Grain:** One row per FILE_ID.

**What It Contains:** Core incident information linking students, cases, locations, and dates. This is the primary table for "how many incidents" questions.
\
```{r}
FactIncident <- df |>
  select(
    FILE_ID,
    CASE_NUMBER,
    SID,
    TYPE,
    STATUS,
    ROLE,
    ASSIGNED_TO,
    ACADEMIC_ORG_CODE,
    INCIDENT_LOCATION,
    INCIDENT_DATE,
    REPORTED_DATE,
    CASE_CREATED_DATE
  )
```

::: {.callout-tip}
## *The Baseline Measure*

**For Decision-Makers:** Incidents are typically the most stable unit for trend analysis. Unlike violations (which depend on charging patterns) or sanctions (which depend on outcome patterns), incidents represent "something happened that was reported."\

**For Analysts:** This table includes all involved parties — respondents, complainants, witnesses, and any other roles present in the data — distinguished by the ROLE field. Most conduct analysis questions are respondent-focused; filter to `ROLE == 'Respondent'` when counting students involved in conduct matters to avoid including complainants and witnesses in those counts. Non-respondent records are retained to support complainant-side analysis if needed. This table connects to most dimensions: DimDate (via `INCIDENT_DATE`), DimHousing (via `INCIDENT_LOCATION`), DimStudent (via `SID`), DimCase (via `FILE_ID`), and DimHearingOfficer (via `ASSIGNED_TO`).
:::

#### FactTimeline

**Grain:** One row per FILE_ID with timeline metrics.

**What It Contains:** Process timing measures (adjusted for pause periods). Supports workflow and efficiency analysis.
\
```{r}
FactTimeline <- df |>
  select(
    FILE_ID,
    INCIDENT_DATE,
    REPORTED_DATE,
    CASE_CREATED_DATE,
    HEARING_DATE,
    reported_to_case,
    case_to_resolution,
    reported_to_resolution
  )
```

::: {.callout-tip}
## *Understanding Process Flow*

**For Decision-Makers:** Use this to identify bottlenecks. If `reported_to_case` is increasing, there may be intake/triage issues. If `case_to_resolution` is increasing, there may be hearing capacity issues.\

**For Analysts:** These metrics have already been adjusted for OMIT_RANGES and floored to zero. Join to FactIncident or DimCase on FILE_ID to add case type, assignment, and other context.
:::

#### FactViolation

**Grain:** One row per alleged violation (charge).

**What It Contains:** Violations and their outcomes. Supports charging pattern and outcome analysis.

**Transformation:** Maxient exports violations in wide format (CHARGE_1, CHARGE_2, etc.). This code reshapes them into long format where each row is one violation and outcome.
\
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Reshape charges from wide to long format
# ─────────────────────────────────────────────────────────────────────────────
charges <- df |>
  select(FILE_ID, INCIDENT_DATE, College, INCIDENT_LOCATION,
         starts_with("CHARGE_")) |>
  pivot_longer(
    cols = starts_with("CHARGE_"),
    names_to = "ChargeCol",
    values_to = "VIOLATION",
    values_drop_na = FALSE
  ) |>
  mutate(
    ChargeIndex = parse_number(ChargeCol)
  )

# ─────────────────────────────────────────────────────────────────────────────
# Reshape findings to match charges
# ─────────────────────────────────────────────────────────────────────────────
findings <- df |>
  select(FILE_ID, starts_with("FINDING_")) |>
  pivot_longer(
    cols = starts_with("FINDING_"),
    names_to = "FindingCol",
    values_to = "OUTCOME",
    values_drop_na = FALSE
  ) |>
  mutate(
    ChargeIndex = parse_number(FindingCol)
  ) |>
  select(FILE_ID, ChargeIndex, OUTCOME)

# ─────────────────────────────────────────────────────────────────────────────
# Join charges with findings and add case type
# ─────────────────────────────────────────────────────────────────────────────
FactViolation <- charges |>
  left_join(findings, by = c("FILE_ID", "ChargeIndex")) |>
  left_join(df |> select(FILE_ID, TYPE), by = "FILE_ID") |>
  
  # Keep only actual violations (not empty rows or excluded types)
  filter(
    !is.na(VIOLATION),
    # These violations are excluded because they point to errors 
    # in charging or case type.
    VIOLATION %nin% c("General Concern", "Aiding and Abetting (Academic)")
  ) |>
  select(
    FILE_ID,
    TYPE,
    INCIDENT_DATE,
    INCIDENT_LOCATION,
    College,
    VIOLATION,
    OUTCOME
  )

# ─────────────────────────────────────────────────────────────────────────────
# Standardize violation names
# ─────────────────────────────────────────────────────────────────────────────
FactViolation <- FactViolation |>
  mutate(VIOLATION = recode(VIOLATION, !!!VIOLATION_REPLACEMENTS, 
                            .default = VIOLATION)) |>
  mutate(OUTCOME = recode(OUTCOME, 
                          "Amnesty - No Hearing" = "Amnesty",
                          'FYI' = 'Founded',
                          'Educational Conference' = 'Educational Response',
                          'Educational Conversation' = 'Educational Response'
                          ),
         OUTCOME = replace_na(OUTCOME, "Unresolved")
         )
```

::: {.callout-tip}
## *Charges vs. Incidents*

**For Decision-Makers:** A single incident might result in three charges (violations). Counting rows in FactViolation tells you how many charges were filed; counting distinct FILE_IDs tells you how many incidents had charges. Both numbers are valid—they answer different questions.\

**For Analysts:** The `ChargeIndex` links each CHARGE_N to its corresponding FINDING_N. The `VIOLATION_REPLACEMENTS` mapping in `config.R` standardizes violation names (e.g., HTML-encoded ampersands, inconsistent naming). If OUTCOME is NULL, the charge is still pending resolution.

Note that `College` is included directly in this fact table as a deliberate denormalization. In a strict star schema, college context would be accessed through a join path rather than stored in the fact table. It is included here because the complexity of the Power BI relationship graph made reliable filter propagation through join paths impractical for college-level violation analysis. Institutions adapting this pipeline should evaluate whether their own model complexity warrants the same approach.
:::

#### FactSanction

**Grain:** One row per sanction applied.

**What It Contains:** Sanctions linked to cases. Supports intervention pattern analysis.

**Transformation:** Similar to violations, sanctions are stored in wide format and reshaped to long format.
\
```{r}
FactSanction <- sanctions |>
  
  # Convert "No" values to NA (only "Yes" values should create rows)
  mutate(across(where(is.character), ~ if_else(.x == "No", NA_character_, .x))) |>
  
  # Reshape from wide to long
  pivot_longer(
    cols = ACAD_ADV_MTG:SUSPENSION,
    names_to = "Sanction",
    values_to = "Value",
    values_drop_na = TRUE
  ) |>
  select(FILE_ID, TYPE, INCIDENT_DATE, College, Sanction)
```

::: {.callout-tip}
## *Intervention Mix*

**For Decision-Makers:** This table shows what actions we're taking. Join to DimSanction to categorize sanctions (educational, restrictive, etc.) and understand your intervention philosophy in practice.\

**For Analysts:** A case can have multiple sanctions — each becomes a separate row. To count "cases with any educational sanction," you'll need: `FactSanction |> inner_join(DimSanction) |> filter(CATEGORY_LABEL == "Educational") |> distinct(FILE_ID) |> count()`. As with FactViolation, `College` is included directly in this table as a deliberate denormalization to support reliable college-level filtering in Power BI.
:::

#### FactRecidivism

**Grain:** One row per academic year with repeat involvement summary statistics.

**What It Contains:** Counts of students found responsible and students found responsible in more than one case within the same academic year.

**Definition:**

***Within-Year Recidivism:*** 
: A student is counted as a recidivist if they were found responsible in **two or more separate cases** (FILE_IDs) within the **same academic year**.

- A student with one case containing multiple responsible findings is **not** a recidivist
- A student with two cases, each with at least one responsible finding, **is** a recidivist
- Recidivism is calculated independently for each academic year (a student can be a recidivist in AY2324 but not in AY2425)

**Example:**

| **Student** | **Cases in AY2324** | **Responsible Findings** | **Recidivist?** |
|---------|-----------------|---------------------|-------------|
| A | 1 case | 3 findings (all in same case) | No |
| B | 2 cases | 1 finding in each case | Yes |
| C | 3 cases | Findings in 2 of 3 cases | Yes |
| D | 1 case | 1 finding | No |
\
```{r}
FactRecidivism <- calculate_recidivism(
  df,
  academic_years = ACADEMIC_YEARS,
  format = "display"
) |>
  select(-Rate)
```

::: {.callout-tip}
## *Using Within-Year Recidivism*

**For Decision-Makers:** This metric answers: "What percentage of students who get in trouble come back to us again within the same academic year?" A rate of 15% means that 15 out of every 100 students found responsible will have another responsible finding before the year ends. This is useful for understanding short-term repeat involvement and the immediate effectiveness of interventions.\

**For Analysts:** The `recidivism()` function counts distinct FILE_IDs with responsible findings per student per academic year. The Rate column is removed here because rates are better calculated in Power BI (allows for dynamic filtering). This measure resets each academic year—a student who was a recidivist in AY2324 starts fresh in AY2425.
:::

#### FactCohortRecidivism

**Grain:** One row per student whose first responsible finding occurred within the analysis period.

**What It Contains:** Cohort-based recidivism tracking showing whether each student ever had a subsequent responsible finding after their first.

**Definition:**

***Cohort-Based Recidivism:***
: A student is assigned to a **cohort** based on the academic year of their **first-ever responsible finding**. They are counted as a recidivist if they have **any subsequent case** with a responsible finding, regardless of how much later it occurs.

- Cohort assignment is based on first responsible finding in the dataset
- "Subsequent" means any case with an incident date after the first case's incident date
- There is no time limit—a student who is found responsible for another violation 3 years later still counts
- Each student appears in exactly one cohort (their first year)

**Example:**

| **Student** | **First Responsible Finding** | **Subsequent Findings** | **Recidivist?** | **Cohort** |
|---------|---------------------------|---------------------|-------------|--------|
| A | AY2122 | None | No | AY2122 |
| B | AY2122 | AY2324 (2 years later) | Yes | AY2122 |
| C | AY2223 | AY2223 (same year) and AY2324 | Yes | AY2223 |
| D | AY2425 | None yet | No* | AY2425 |

*Student D may have a violation in the future—their cohort has limited follow-up time.\

**How This Differs from FactRecidivism:**

| **Aspect** | **FactRecidivism** | **FactCohortRecidivism** |
|--------|----------------|----------------------|
| **Question answered** | "How many repeat in the same year?" | "How many ever come back?" |
| **Time window** | Single academic year | All time after first finding |
| **Grain** | One row per academic year | One row per student |
| **A student can appear** | In multiple years | In exactly one cohort |
| **Comparable to** | Annual repeat rate | Criminal justice recidivism |
\
```{r}
FactCohortRecidivism <- calculate_cohort_recidivism(
  df,
  cohort_years = ACADEMIC_YEARS
)
```

::: {.callout-tip}
## *Understanding Cohort Recidivism*

**For Decision-Makers:** This table answers: Of the students who first got in trouble in a given year, how many ever got in trouble again? This is how recidivism is reported in criminal justice and other fields, making it useful for benchmarking and for evaluating whether interventions have lasting impact.\

***Important caveat:*** Recent cohorts (e.g., AY2425) will show lower recidivism rates simply because less time has passed—not because interventions improved. Always compare cohorts with similar follow-up periods, or note the follow-up time when reporting.\

**For Analysts:** Key measures you can calculate in Power BI: Recidivism Rate and Cohort Size.\

Join to DimStudent on SID to disaggregate by demographics. Join to DimDate on First_Incident_Date for time-based filtering.
:::

**Key Fields:**

| **Field** | **Description** |
|-------|-------------|
| SID | Student identifier (anonymized) |
| Cohort_Year | Academic year of student's first responsible finding |
| First_Incident_Date | Date of first responsible finding |
| Total_Responsible_Cases | Total number of cases with responsible findings |
| Repeat_Count | Number of cases after the first (0 if no recidivism) |
| Is_Recidivist | TRUE if student had any subsequent responsible finding |
| Recidivism_Category | Intensity grouping: No Recidivism, One Repeat, Two Repeats, Three+ Repeats |
| Second_Incident_Date | Date of second responsible finding (NULL if no recidivism) |
| Second_Case_Year | Academic year of second responsible finding |
| Days_to_Recidivism | Days between first and second responsible finding (NULL if no recidivism) |
\
**Sample Analysis Questions This Enables:**

| **Question** | **How to Answer** |
|----------|---------------|
| What's our overall recidivism rate? | % where Is_Recidivist = TRUE |
| Which cohort year has the highest recidivism? | Rate by Cohort_Year (with follow-up time caveat) |
| How quickly do students recidivate? | Median/distribution of Days_to_Recidivism |
| Do certain student groups have higher recidivism? | Join to DimStudent, compare rates |
| Are students with more severe first sanctions less likely to recidivate? | Join first case to FactSanction, compare rates |
| How intense is repeat involvement? | Distribution of Recidivism_Category |
| What % of recidivists have 3+ cases? | Filter to Is_Recidivist = TRUE, count Three+ Repeats |
| What's the average number of repeats among recidivists? | Average Repeat_Count where Is_Recidivist = TRUE |

#### FactHousingCensus

**Grain:** One row per building per term (Fall/Spring only).

**What It Contains:** Violation counts, census (resident) counts, and calculated violation rates per 100 residents. Supports rate-based housing analysis.

**Why Fall/Spring Only:** Summer census is typically too small for meaningful rate calculations.

**A Note on Census Data Format:** The reshape code below reflects one institution's source file structure. If your census data is formatted differently, transform it to long format — one row per building per term — before Step 3. The join logic and rate calculation in Step 3 are institution-agnostic once the data is in that shape.
\
```{r message=FALSE}
# ─────────────────────────────────────────────────────────────────────────────
# Step 1: Count violations by building and term
# ─────────────────────────────────────────────────────────────────────────────
violation_counts <- FactViolation |>
  left_join(
    DimDate |> select(Date, Academic_Year = `Academic Year`, Semester),
    by = c("INCIDENT_DATE" = "Date")
  ) |>
  filter(Semester %in% c("FA", "SP")) |>
  group_by(INCIDENT_LOCATION, Academic_Year, Semester) |>
  summarise(VIOLATIONS = n(), .groups = "drop")

# ─────────────────────────────────────────────────────────────────────────────
# Step 2: Load and reshape census data
# ─────────────────────────────────────────────────────────────────────────────
census_long <- read_csv(PATHS$imports$housing_census) |>
  rename_with(str_to_upper) |>
  pivot_longer(
    cols = -BUILDING,
    names_to = "Semester",
    values_to = "CENSUS"
  ) |>
  rename(INCIDENT_LOCATION = BUILDING) |>
  mutate(INCIDENT_LOCATION = fix_nbsp(INCIDENT_LOCATION)) |>
  mutate(
    # Map semester codes to academic years
    Academic_Year = case_when(
      Semester %in% c("FA17", "SP18") ~ "AY1718",
      Semester %in% c("FA18", "SP19") ~ "AY1819",
      Semester %in% c("FA19", "SP20") ~ "AY1920",
      Semester %in% c("FA20", "SP21") ~ "AY2021",
      Semester %in% c("FA21", "SP22") ~ "AY2122",
      Semester %in% c("FA22", "SP23") ~ "AY2223",
      Semester %in% c("FA23", "SP24") ~ "AY2324",
      Semester %in% c("FA24", "SP25") ~ "AY2425",
      TRUE ~ "AY2526"
    ),
    # Standardize semester to FA/SP format
    Semester = str_sub(Semester, 1, 2)
  ) |>
  select(INCIDENT_LOCATION, Academic_Year, Semester, CENSUS)

# ─────────────────────────────────────────────────────────────────────────────
# Step 3: Combine census with violations and calculate rates
# ─────────────────────────────────────────────────────────────────────────────
FactHousingCensus <- census_long |>
  left_join(
    violation_counts,
    by = c("INCIDENT_LOCATION", "Academic_Year", "Semester")
  ) |>
  left_join(
    DimAcademicTerm,
    by = c("Academic_Year", "Semester")
  ) |>
  mutate(
    # Replace NA with 0 for clean calculations
    CENSUS     = replace_na(CENSUS, 0),
    VIOLATIONS = replace_na(VIOLATIONS, 0),
    
    # Calculate rate per 100 residents (NULL if no census)
    Violation_Rate = if_else(
      CENSUS > 0,
      round((VIOLATIONS / CENSUS) * 100, 2),
      NA_real_
    )
  ) |>
  select(
    AcademicTermKey,
    INCIDENT_LOCATION,
    Academic_Year,
    Semester,
    CENSUS,
    VIOLATIONS,
    Violation_Rate
  ) |>
  # Exclude temporary housing
  filter(INCIDENT_LOCATION != "MainStay Suites")
```

::: {.callout-tip}
## *Fair Comparisons Across Buildings*

**For Decision-Makers:** A hall with 50 violations and 800 residents (6.25 per 100) is very different from a hall with 50 violations and 200 residents (25 per 100). This table pre-calculates rates so you can compare fairly.\

**For Analysts:** When CENSUS is 0 (building closed or not yet open), Violation_Rate is NULL—not zero. This prevents misleading rate calculations. If a building has violations but no census record, VIOLATIONS will be populated but CENSUS and Violation_Rate will be 0/NULL—investigate the census data.
:::

#### FactResolution

**Grain:** One row per resolved FILE_ID.

**What It Contains:** Resolution method and hearing date for cases that reached a formal resolution. Supports analysis of how cases are resolved and workload distribution across hearing officers.

| **Key Fields** | **Description** |
|------------|-------------|
| FILE_ID | Unique case identifier (anonymized) |
| HEARING_DATE | Date the case was resolved |
| Resolution | Method of resolution (e.g., Administrative Hearing, Informal Resolution) |
| ASSIGNED_TO | Hearing officer assignment code |

**Note:** This table excludes "Unheard" cases and cases without a recorded hearing date. Use this table when your question is specifically about resolved cases and resolution methods rather than all incidents. `ASSIGNED_TO` is included directly in this table as a deliberate denormalization to support reliable visualization of resolutions by hearing officer across months and ISO weeks without requiring additional join paths in Power BI.
\
```{r}
FactResolution <- df |>
  select(FILE_ID,
         HEARING_DATE,
         Resolution,
         ASSIGNED_TO) |>
  filter(Resolution != 'Unheard',
         !is.na(HEARING_DATE))
```

#### FactTags and DimTag

**Grain (FactTag):** One row per tag applied to a case.

**What They Contain:** Institution-specific case tags that flag operationally significant conditions — such as dismissed cases, health and safety inspections, or hospital transports. `DimTag` provides a controlled list of all possible tag values.

| **Key Fields** | **Description** |
|------------|-------------|
| FILE_ID | Unique case identifier (anonymized) |
| Tag | The tag applied to the case |

**Note:** Tags are institution-specific and reflect their Maxient configuration. Institutions adapting this pipeline would need to update tag definitions and field names to match their own configuration. Tags are useful for operational filtering — for example, identifying dismissed cases or flagging cases involving hospital transport — but are not intended for trend analysis.

::: {.callout-tip}
## *Using Tags in Analysis*

**For Decision-Makers:** Tags flag cases that may need separate handling in reports. For example, filtering out dismissed cases before calculating outcome rates ensures you're analyzing cases that actually went through the process.

**For Analysts:** `FactTag` is in long format — one row per tag per case. A single case can carry multiple tags. Join to `DimTag` on the Tag field for filtering. When building dashboards, consider whether tagged cases should be included or excluded from core metrics, and document that decision explicitly.
:::
\
```{r}
FactTag <- tags |>
  pivot_longer(
    cols = -FILE_ID,
    names_to = "Tag",
    values_to = "Flag"
  ) |>
  filter(!is.na(Flag)) |>
  mutate(Tag = str_remove(Tag, "^Tag: "),
         Tag = str_replace_all(Tag, "&amp;", "and")
         ) |>
  select(FILE_ID, Tag)

DimTag <- FactTag |>
  distinct(Tag) |>
  arrange(Tag)
```

## Export Star Schema

The final step exports all dimension and fact tables into a single Excel workbook. Power BI imports this workbook, with each worksheet becoming a table.
\
```{r}

output_list <- list(
  # ───────────────────────────────────────────────────────────────────────
  # Dimension Tables
  # ───────────────────────────────────────────────────────────────────────
  DimStudent        = DimStudent,
  DimCase           = DimCase,
  DimCollege        = DimCollege,
  DimHousing        = DimHousing,
  DimHousingYear    = DimHousingYear,
  DimSanction       = DimSanction,
  DimDate           = DimDate,
  DimAcademicYear   = DimAcademicYear,
  DimHearingOfficer = DimHearingOfficer,
  DimAcademicTerm   = DimAcademicTerm,
  DimViolation      = DimViolation,
  DimTag            = DimTag,
  DimAcademicPlans  = DimAcademicPlans,
    
  # ───────────────────────────────────────────────────────────────────────
  # Fact Tables
  # ───────────────────────────────────────────────────────────────────────
  FactIncident         = FactIncident,
  FactTimeline         = FactTimeline,
  FactViolation        = FactViolation,
  FactSanction         = FactSanction,
  FactRecidivism       = FactRecidivism,
  FactCohortRecidivism = FactCohortRecidivism,
  FactDeadline         = FactDeadline,
  FactHousingCensus    = FactHousingCensus,
  FactResolution       = FactResolution,
  FactTag              = FactTag
  )

if (ANONYMIZE) {
  write_xlsx(output_list, PATHS$export_local)
}
```

::: {.callout-tip}
## *Single Source of Truth*

**For Decision-Makers:** This workbook is the single source for all conduct dashboards. When it's refreshed, all dashboards update together using consistent definitions. Note that the export file is only created when the `ANONYMIZE` flag is set to `TRUE` at the top of the pipeline. This is an intentional safeguard: it prevents identifiable data from being written to disk unless anonymization has been applied.

**For Analysts:** After import into Power BI, establish relationships:

- Fact tables → DimDate on date fields
- Fact tables → DimCase on FILE_ID
- FactIncident → DimStudent on SID
- FactIncident/FactViolation → DimHousing on INCIDENT_LOCATION
- FactSanction → DimSanction on Sanction
- DimCase → DimHearingOfficer on ASSIGNED_TO

The export path is defined in config.R (`PATHS$export`), making it easy to change output locations without modifying the pipeline code.
:::

\newpage

# Power BI Analytical Layer

This section documents the analytical layer that sits on top of the exported star-schema workbook in Power BI. It records the relationships established between tables and summarizes the DAX measures used in the accompanying dashboards. This is intended as a technical reference for transparency, reproducibility, and knowledge transfer. Those focused solely on methodology or outcomes may choose to skip it.

## Table Relationships

The table below lists the active relationships that define the Power BI semantic model.  
These relationships determine how filters propagate across the star schema and ensure that measures aggregate consistently across visuals.

| **From** | **To** | **Cardinality** | **Filter** |
|----------|----------|------|------|
| FactIncident[INCIDENT_DATE] | DimDate[Date] | Many-to-One | Single |
| FactViolation[INCIDENT_DATE] | DimDate[Date] | Many-to-One | Single |
| FactSanction[INCIDENT_DATE] | DimDate[Date] | Many-to-One | Single |
| FactTimeline[INCIDENT_DATE] | DimDate[Date] | Many-to-One | Single |
| FactDeadline[DEADLINE] | DimDate[Date] | Many-to-One | Single |
| FactIncident[SID] | DimStudent[SID] | Many-to-One | Single |
| FactCohortRecidivism[SID] | DimStudent[SID] | One-to-One | Both |
| FactIncident[FILE_ID] | DimCase[FILE_ID] | Many-to-One | Single |
| FactViolation[FILE_ID] | DimCase[FILE_ID] | Many-to-One | Single |
| FactSanction[FILE_ID] | DimCase[FILE_ID] | Many-to-One | Single |
| FactTimeline[FILE_ID] | DimCase[FILE_ID] | Many-to-One | Single |
| FactDeadline[FILE_ID] | DimCase[FILE_ID] | Many-to-One | Single |
| FactIncident[College] | DimCollege[College] | Many-to-One | Single |
| FactViolation[College] | DimCollege[College] | Many-to-One | Single |
| FactSanction[College] | DimCollege[College] | Many-to-One | Single |
| FactIncident[INCIDENT_LOCATION] | DimHousing[INCIDENT_LOCATION] | Many-to-One | Single |
| FactViolation[INCIDENT_LOCATION] | DimHousing[INCIDENT_LOCATION] | Many-to-One | Single |
| FactHousingCensus[INCIDENT_LOCATION] | DimHousing[INCIDENT_LOCATION] | Many-to-One | Single |
| FactHousingCensus[INCIDENT_LOCATION] | DimHousingYear[INCIDENT_LOCATION] | Many-to-One | Single |
| FactHousingCensus[AcademicTermKey] | DimAcademicTerm[AcademicTermKey] | Many-to-One | Single |
| FactRecidivism[ACADEMIC_YEAR] | DimAcademicYear[ACADEMIC_YEAR] | Many-to-One | Single |
| DimDate[AcademicTermKey] | DimAcademicTerm[AcademicTermKey] | Many-to-One | Single |
| DimDate[Academic Year] | DimAcademicYear[ACADEMIC_YEAR] | Many-to-One | Single |
| DimHousingYear[ACADEMIC_YEAR] | DimAcademicYear[ACADEMIC_YEAR] | Many-to-One | Single |
| DimCase[ASSIGNED_TO] | DimHearingOfficer[ASSIGNED_TO] | Many-to-One | Single |
| FactSanction[SANCTION] | DimSanction[SANCTION] | Many-to-One | Single |
| FactViolation[VIOLATION] | DimViolation[VIOLATION] | Many-to-One | Single |

**In summary:** All fact tables filter into dimension tables using single-direction, many-to-one relationships.\

`DimDate` is designated as the model’s official Date Table in Power BI, and all time-based calculations and calendar filtering are anchored to `DimDate[Date]`. Dimension-to-dimension relationships (such as Date → Academic Year and Date → Academic Term) support hierarchical filtering across dashboards while preserving a clear star-schema structure.

## DAX Measures

The following measures support the dashboards built on this model. Full DAX expressions are provided in the appendix for technical reference.\

| **Measure** | **Purpose** |
|-------|--------------|
| **Total Cases** | Counts distinct incident cases |
| **Total Charges** | Counts all violation records |
| **Total Sanctions** | Counts all sanctions issued |
| **Distinct Cases Charged** | Counts unique cases with at least one charge |
| **Distinct Cases with Sanctions** | Counts unique cases receiving sanctions |
| **Sanctions per Case** | Average number of sanctions per sanctioned case |
| **Violation Count** | Counts all violations |
| **Violation Count (AOD FA/SP Only)** | Counts alcohol/drug violations during Fall/Spring |
| **Responsible Violations** | Counts violations with a “Responsible” outcome |
| **Responsible Rate** | Percent of violations resulting in responsibility |
| **Not Responsible Rate** | Percent of violations resulting in non-responsibility |
| **Unresolved Rate** | Percent of violations without a final finding |
| **Other Finding Rate** | Percent of violations with alternative outcomes |
| **Educational Sanctions** | Counts sanctions categorized as educational |
| **Restrictive Sanctions** | Counts sanctions with restrictive impact |
| **Academic Sanctions** | Counts sanctions affecting academic standing |
| **Housing Sanctions** | Counts housing-related sanctions |
| **Removal Sanctions** | Counts sanctions involving removal/separation |
| **Max Severity** | Highest severity level assigned in context |
| **Avg Severity per Case (by Incident Term)** | Average maximum sanction severity per case |
| **Total Census** | Sum of housing population counts |
| **Total Violations** | Sum of housing violation counts |
| **Violations per 100 Residents (Term Safe)** | Normalized housing violation rate |
| **Alcohol Rate per 100 Residents** | Alcohol violations normalized by housing population |
| **Drug Rate per 100 Residents** | Drug violations normalized by housing population |
| **Census Denominator (Sum of Building-Year Max)** | Stable denominator for rate calculations |
| **Avg Annual Census (Max FA/SP, Active Years Only)** | Average active annual housing population |
| **reported_to_case_creation** | Average time from report to case creation |
| **case_creation_to_resolution** | Average time from case creation to resolution |
| **report_to_resolution** | Average time from report to resolution |
| **RED Ed Resp Count** | Counts educational response incidents |
| **Time Saved (Minutes)** | Estimated minutes saved via educational responses |
| **Time Saved (Hours)** | Estimated hours saved via educational responses |
| **Recidivism Rate** | Percent of students with repeat responsible findings |
| **Total Holds** | Counts cases with holds in place |
| **Total Active Holds** | Counts unresolved active holds |
| **Hold Rate** | Percentage of cases with holds |
| **Total Overdue** | Counts overdue deadlines |
| **Overdue Rate** | Percentage of cases overdue |

\newpage

# R Computing Environment
```{r echo=FALSE}
si <- capture.output(
  suppressWarnings(
    sessioninfo::session_info(pkgs = "attached")
  )
)

cat(
  gsub("([A-Z]:)?[/\\\\][^[:space:]]+", "<path>", si),
  sep = "\n"
)
```

\newpage

# About the Author

Josh Moermond serves as the Assistant Director for Residential Conduct and Assessment at the University of Cincinnati, where his work sits at the intersection of student conduct administration and institutional assessment. He has spent nearly a decade in student affairs, with experience spanning conduct case management, data systems administration, and assessment practice.

Moermond's approach to analytics is grounded in practitioner experience. Having navigated the limitations of conduct data firsthand, he developed the pipeline described in this document to connect operational case management data with the institutional context needed to make it analytically useful. His work focuses on building systems that are reproducible, privacy-conscious, and accessible to the colleagues who need to use them.

In addition to his assessment and data work, Moermond maintains expertise in threat assessment and behavioral intervention, holding multiple certifications from the National Association for Behavioral Intervention and Threat Assessment (NABITA).

Moermond holds a Master of Arts in Educational Leadership from Eastern Michigan University and a Graduate Certificate in Data Analytics from the University of Cincinnati.

He can be reached at [moermondsahe@gmail.com](mailto:moermondsahe@gmail.com) or via [LinkedIn](https://www.linkedin.com/in/jmoermond).

\newpage

# Use of Generative AI

Portions of this document were developed with the assistance of **generative AI tools.**

Generative AI was used as a writing and technical support aid, including:

- Refining language for clarity and accessibility
- Improving organization and flow of explanatory text
- Assisting with code errors, readability, formatting, custom functions, and documentation
- Helping draft example explanations, tables, and narrative framing

All analytic decisions, data transformations, methodological choices, and interpretations were designed, reviewed, and validated by the author. The underlying logic of the data model, transformation pipeline, and reporting framework reflects professional judgment grounded in student conduct practice and assessment standards.

No confidential, proprietary, or student-level data was shared with generative AI systems as part of this work.

The author retains full responsibility for the content, accuracy, and conclusions presented.

\newpage

# References

::: {#refs}
:::

\appendix

# Custom Functions (conduct_funcs.R)

This file contains reusable functions for data transformation, anonymization, and analysis.
\
```{r}
#| echo: false
#| output: asis
#| warning: false

cat("\\footnotesize\n")
cat("```r\n")
cat(readLines("~/GitHub/conduct-analytics/shared/R/conduct_funcs.R"), sep = "\n")
cat("\n```\n")
cat("\\normalsize\n")
```

\newpage

::: {.landscape}
# Configuration File (config.R)

This file contains all constants, mappings, and settings used by the pipeline. Update this file to add new academic years, modify location mappings, or adjust pause periods.
\
```{r}
#| echo: false
#| output: asis
#| code-overflow: wrap

config_lines <- readLines("~/GitHub/conduct-analytics/shared/R/config.R")
config_redacted <- gsub("([A-Z]:)?[/\\\\][^[:space:]]+", "<path>", config_lines)

cat("```r\n")
cat(config_redacted, sep = "\n")
cat("\n```")
```
:::
